---
title: "Robots and Taxes"
subtitle: "EC 350: Labor Economics"
author: "[Kyle Raze](https://kyleraze.com)"
date: "Winter 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
      ratio: "16:9"
---

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(ggthemes, xaringanExtra, ggdag, viridis, knitr, dslabs, gapminder, extrafont, Ecdat, wooldridge, tidyverse, magrittr, janitor, kableExtra)
# Define colors
red_pink <- "#e64173"
met_slate <- "#272822" # metropolis font color 
purple <- "#9370DB"
green <- "#007935"
light_green <- "#7DBA97"
orange <- "#FD5F00"
turquoise <- "#44C1C4"
# Notes directory
dir_slides <- "~/GitHub/Intro_Labor_S21/Lectures/05-Workers/"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  #dpi = 300,
  #cache = T,
  warning = F,
  message = F
)  
theme_simple <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
theme_market <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.title.x = element_text(hjust = 1, size = 17),
  axis.title.y = element_text(hjust = 1, angle = 0, size = 17),
  # axis.text.x = element_text(size = 12),
  # axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
theme_gif <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
wrapper <- function(x, ...) paste(strwrap(x, ...), collapse = "\n")
shift_axis <- function(p, y=0){
  g <- ggplotGrob(p)
  dummy <- data.frame(y=y)
  ax <- g[["grobs"]][g$layout$name == "axis-b"][[1]]
  p + annotation_custom(grid::grobTree(ax, vp = grid::viewport(y=1, height=sum(ax$height))), 
                        ymax=y, ymin=y) +
    geom_hline(aes(yintercept=y), data = dummy, size = 0.5, color = met_slate) +
    theme(axis.text.x = element_blank(), 
          axis.ticks.x = element_blank())
}

xaringanExtra::use_scribble()
```

# Agenda

1. Factor demand
    - Hiring in the short run (review)
    - Hiring in the long run
    - Labor demand curve
3. Robot tax, featuring Bill Gates

---
class: inverse, middle

# Factor demand

---
# Short run *vs.* long run

## **Short run**

> The time span over which a business can adjust some inputs (*e.g.,* labor), but cannot adjust others (*e.g.,* capital).

In the short run, we will assume that the level of employment **E** can vary, but capital **K** is fixed at an initial level **K.sub[0]**.

- **Example:** A shop foreman can hire or fire workers or adjust hours, but they are unable to expand the factory by adding assembly lines, heavy machinery, or a new building.

---
# Short run *vs.* long run

## **Long run**

> The time span over which a business can adjust all inputs.

In the long-run, we will assume that both the level of employment **E** and capital **K** can vary.

- **Example:** An office manager can hire or fire workers, adjust hours, buy or sell desks and computers, or lease new office space.

---
# Hiring in the short run

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
wage <- 1000
vmp <- function(x) (- 1.5 * x ^ 2 + 10 * x) * 150
vap <- function(x) (- 0.5 * x ^ 2 + 5 * x) * 150 
inflection <- uniroot(function(x) 10 - 3 * x, c(0, 7))$root
break_even <- uniroot(function(x) (- 1.5 * x ^ 2 + 10 * x) * 150 - (- 0.5 * x ^ 2 + 5 * x) * 150, c(inflection, 7))$ root
optimal_e <- uniroot(function(x) (- 1.5 * x ^ 2 + 10 * x) * 150 - wage, c(inflection, 7))$root

ggplot() +
  stat_function(
    data = tibble(x = 0:10),
    xlim = c(0, 7),
    fun = vap,
    aes(x = x),
    color = purple,
    size = 1,
  ) +
  stat_function(
    data = tibble(x = 0:10),
    xlim = c(0, 7),
    fun = vmp,
    aes(x = x),
    color = purple,
    size = 1,
  ) +
  scale_x_continuous(limits = c(0, 7.5), expand = c(0, 0), breaks = seq(0, 7, 1)) +
  scale_y_continuous(limits = c(0, 2750), expand = c(0, 0), breaks = seq(0, 2500, 500), labels = scales::dollar) +
  labs(x = "Number of workers (E)", y = "Dollars") +
  theme_simple +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10))) +
  geom_hline(yintercept = wage, size = 1, color = green) +
  annotate(geom = "segment", x = optimal_e, y = 0, xend = optimal_e, yend = wage,
           linetype = "dashed", color = met_slate, alpha = 1) +
  annotate(geom = "point", x = optimal_e, y = wage, size = 2, color = met_slate, alpha = 1) +
  annotate("label", label = "VAP", x = 4, y = vap(4), color = purple, family = "Fira Sans", size = 4) +
  annotate("label", label = "VMP = MRP", x = 4, y = vmp(4), color = purple, family = "Fira Sans", size = 4) +
  annotate("label", label = "w", x = 4, y = 1000, color = green, family = "Fira Sans", size = 4)
```
]
.pull-right[

## **Profit maximization**

An employer maximizes profit by hiring $E^*$ workers where $w = \text{MRP}_E$ and $\text{MRP}_E$ is decreasing.

]

---
# Hiring in the short run

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
ggplot() +
  stat_function(
    data = tibble(x = 0:10),
    xlim = c(0, 7),
    fun = vap,
    aes(x = x),
    color = purple,
    size = 1,
    alpha = 0.5
  ) +
  stat_function(
    data = tibble(x = 0:10),
    xlim = c(0, 7),
    fun = vmp,
    aes(x = x),
    color = purple,
    size = 1,
    alpha = 0.5
  ) +
  stat_function(
    data = tibble(x = 0:10),
    xlim = c(break_even, 7),
    fun = vmp,
    aes(x = x),
    color = purple,
    size = 2,
  ) +
  scale_x_continuous(limits = c(0, 7.5), expand = c(0, 0), breaks = seq(0, 7, 1)) +
  scale_y_continuous(limits = c(0, 2750), expand = c(0, 0), breaks = seq(0, 2500, 500), labels = scales::dollar) +
  labs(x = "Number of workers (E)", y = "Wage") +
  theme_simple +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10))) +
  annotate("label", label = "VAP", x = 4, y = vap(4), color = purple, family = "Fira Sans", size = 4, alpha = 1) +
  annotate("label", label = "VMP = MRP", x = 4, y = vmp(4), color = purple, family = "Fira Sans", size = 4, alpha = 1) +
annotate("label", label = "D", x = 6, y = vmp(6), color = purple, family = "Fira Sans", size = 5)
```
]
.pull-right[

## .hi-purple[Labor demand]

The portion of the MRP curve below the VAP curve traces out the .hi-purple[short-run labor demand] curve.

- Describes how an employer adjusts employment as the market wage changes, holding other inputs constant.
- **Downward sloping:** An employer wants to reduce staffing as the wage increases, *all else equal*.

]

---
# Hiring in the long run

## **Profit maximization**

In the long run, employers have the flexibility to adjust both labor and capital.

**Q:** How does the intuition for long-run factor demand compare to the intuition for the short run?

--

**A:** Employers will still make decisions ***at the margin*!**

- The underlying model is more elaborate and **our time is scarce**, so we will forgo a full derivation of the long-run profit maximization conditions.

---
# Hiring in the long run

## **Profit maximization**

As in the short run, an employer will choose the profit-maximizing level of employment $E^*$ such that $$\begin{align} w &= \text{MRP}_E \\ &= p \times \text{MP}_E \end{align}$$

--

- The employer will keep hiring until the **marginal cost** of the last worker **equals** the **marginal benefit** of the last worker.


--

Likewise, an employer will choose the profit-maximizing quantity of capital $K^*$ such that $$\begin{align} r &= \text{MRP}_K \\ &= p \times \text{MP}_K \end{align}$$

--

- The employer will keep purchasing capital until the **marginal cost** of the last unit of capital **equals** the **marginal benefit** of the last unit of capital.

---
# Hiring in the long run

## **Intuition?**

At the optimal, profit-maximizing bundle of labor and capital, we have $$\begin{align} w = p \times \text{MP}_E && r = p \times \text{MP}_K \end{align}$$

--

Dividing by marginal product, we obtain $$\begin{align} p = \dfrac{w}{\text{MP}_E} && p = \dfrac{r}{\text{MP}_K} \end{align}$$

--

Now we can equate both conditions to obtain the **long-run profit maximization condition** $$\begin{align} \dfrac{w}{\text{MP}_E} = \dfrac{r}{\text{MP}_K} \end{align}$$

---
# Hiring in the long run

## **Intuition?**

**Q:** What can we learn from the long-run profit maximization condition? $$\begin{align} \dfrac{w}{\text{MP}_E} = \dfrac{r}{\text{MP}_K} \end{align}$$

- $\dfrac{w}{\text{MP}_E}$ represents the cost of producing one more unit of output using labor.
- $\dfrac{r}{\text{MP}_K}$ represents the cost of producing one more unit of output using capital.


--

**A:** At the profit-maximizing bundle, it would **not** be **cheaper to change** the mix of inputs.

- If it were cheaper to use relatively more labor than capital (or vice versa), then a **profit-maximizing employer would have already made the adjustment** (in the long run).

---
# Hiring in the long run

## **Intuition?**

Suppose that $w = 10$, $\text{MP}_E = 15$, $r = 5$, and $\text{MP}_K = 10$. Then $$\dfrac{w}{\text{MP}_E} = \dfrac{r}{\text{MP}_K} \implies \dfrac{10}{15} = \dfrac{5}{10} \implies \dfrac{2}{3} \neq \dfrac{1}{2}$$

--

This employer is not profit maximizing!

- The cost of producing one more unit of output using labor exceeds the cost of producing one more unit using capital.
- It would be **more profitable** to use (relatively) **more capital** or **less labor**!

---
# Hiring in the long run

**Q:** How does an employer respond to an increase in the market wage?

--

- **A.sub[1]:** The employer will respond by hiring fewer workers.

--


- **A.sub[2]:** The employer will adjust the level of capital, but the direction is theoretically ambiguous.
    - When the number of workers decreases, there are fewer people on each machine, which can reduce $\text{MP}_K$.
    - The direction of the response will depends on the **scale** and **substitution** effects.

---
# Hiring in the long run

## **Scale effect**

Other things being equal, a **decrease** in the **price** of an input will **increase** the **quantity demanded** of that input.

- If the cost of production decreases, the employer will want to "scale up" production of the output good.
- Conversely, if the cost of production increases, the employer will "scale back" production.
--

- Analogous to the wealth effect for a worker..super[.hi-pink[<span>&#8224;</span>]]

.footnote[.super[.hi-pink[<span>&#8224;</span>]] We assume that labor and capital are "normal" inputs&mdash;production increases as the amount of labor and capital increase.]


---
# Hiring in the long run

## **Substitution effect**

Other things being equal, if the price of an input increases, **demand for the other input increases.**

- If labor becomes **relatively more expensive** than capital, then the employer will want to **substitute away** from labor and toward capital.
- If labor becomes **relatively cheaper** than capital, then the employer will want to **substitute toward** labor and away from capital.
--

- Analogous to the substitution effect for the worker.

---
# Hiring in the long run

## **Scale and substitution effects**

**Q:** How would a employer respond to an increase in the market wage?


|                     | Scale effect | Substitution effect |
|:-------------------:|:------------:|:-------------------:|
| $\Delta$ in labor   | $-$          | $-$                 |
| $\Delta$ in capital | $-$          | $+$                 |

--

**A.sub[K]:** For capital, it depends.

- If the scale effect dominates the substitution effect, then capital will eventually decrease.

--

**A.sub[E]:** For labor, the effect is unambiguous.

- The scale effect and substitution effect will move in the same direction for the input that undergoes a change in price.

---
# Hiring in the long run

## **Scale and substitution effects**

**Q:** What determines whether the scale or substitution effect dominates?

--

**A:** Whether labor and capital are .hi-pink[substitutes] or .hi-purple[complements].

--

- .hi-pink[Substitutes:] Inputs used in place of one another.
    - Self-checkout kiosk *vs.* cashier
    - Tax prep software *vs.* accountant
    - Robot *vs.* low-skill worker?
- .hi-purple[Complements:] Inputs used together.
    - Carpenter and hammer
    - Mail carrier and mail truck
    - Robot and high-skill worker?


---
# Hiring in the long run

### .hi-pink[Substitutes]

> Inputs used in place of one another.

Two inputs are said to be substitutes if the price of one input changes the demand of the other input .hi-pink[in the same direction.]

- Substitution effect outweighs the scale effect.

--

### .hi-purple[Complements]

> Inputs used together.

Two inputs are said to be complements if the price of one input changes the demand of the other input .hi-purple[in the opposite direction.]

- Scale effect outweighs the substitution effect.

---
# Hiring in the long run

### **Cross-elasticity of factor demand**

> A unit-free measure of the responsiveness of demand for one input to a change in the price of the other. 

**Labor responsiveness** to a change in the price of capital: $$\eta = \dfrac{\%\Delta E}{\%\Delta r} = \dfrac{(E_2 - E_1) / E_1}{(r_2 - r_1) / r_1}$$

**Capital responsiveness** to a change in the price of labor: $$\eta = \dfrac{\%\Delta K}{\%\Delta w} = \dfrac{(K_2 - K_1) / K_1}{(w_2 - w_1) / w_1}$$

--

$\eta > 0$ .mono[-->] substitutes .mono[-->] substitution effect dominates. <br>
$\eta < 0$ .mono[-->] complements .mono[-->] scale effect dominates.

---
# Hiring in the long run

### **Cross-elasticity of factor demand**

> A unit-free measure of the responsiveness of demand for one input to a change in the price of the other. 

**Q:** What is the cross-elasticity of factor demand implied by the data below?

|  Month   | Rental rate of capital | Hourly wage | Units of capital | Hours of labor |
|:--------:|:----------------------:|:-----------:|:----------------:|:--------------:|
| January  | 20                     | 15          | 5                | 60             |
| February | 25                     | 15          | 1                | 80             |

--

**A:** Roughly 1.33.

--

$$\eta = \dfrac{\%\Delta E}{\%\Delta r} = \dfrac{(E_2 - E_1) / E_1}{(r_2 - r_1) / r_1} = \dfrac{(80-60)/60}{(25-20)/20} = \dfrac{1/3}{1/4} = \dfrac{4}{3} \approx 1.33$$

---
# Hiring in the long run

### **Cross-elasticity of factor demand**

> A unit-free measure of the responsiveness of demand for one input to a change in the price of the other. 

**Q:** How do we interpret the cross-elasticity of factor demand we just calculated? What does 1.33 tell us?

--

**A.sub[1]:** A 1-percent increase in the price of capital .mono[-->] 1.33-percent increase in labor demanded.

--

**A.sub[2]:** Labor and capital are substitutes.

- Substitution effect .mono[>] scale effect

---
# Market labor demand curve

.pull-left[
```{r, dev = "svg", echo = F, fig.height = 5.5, fig.width = 5.5}
sr_demand <- function(x) 50 - 0.75 * x
lr_demand <- function(x) 25 - 0.25 * x
  
plot <- ggplot() + 
  stat_function(
    data = tibble(x = 0:100),
    xlim = c(0, 100),
    aes(x = x),
    fun = lr_demand,
    color = purple,
    size = 1
  ) +
  annotate(
    "label",
    label = expression(D[LR]),
    x = 10,
    y = lr_demand(x = 10),
    color = purple,
    family = "Fira Sans",
    size = 5
  ) +
  stat_function(
    data = tibble(x = 0:100),
    xlim = c(0, 100),
    aes(x = x),
    fun = sr_demand,
    color = purple,
    size = 1
  ) +
  annotate(
    "label",
    label = expression(D[SR]),
    x = 10,
    y = sr_demand(x = 10),
    color = purple,
    family = "Fira Sans",
    size = 5
  ) +
  # appearance
  labs(x = "Units of labor (E)", y = "Wage (w)") +
  scale_x_continuous(expand = c(0, 0), breaks = seq(0, 100, 20), limits = c(0, 110)) +
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 50, 10), labels = scales::dollar, limits = c(0, 55)) +
  theme_simple +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))

plot
```
]
.pull-right[

**Q:** Why is the short-run labor demand curve (.hi-purple[D.sub[SR]]) steeper than the short-run labor demand curve (.hi-purple[D.sub[LR]])?


]

---
count: false
# Market labor demand curve

.pull-left[
```{r, dev = "svg", echo = F, fig.height = 5.5, fig.width = 5.5}
plot
```
]
.pull-right[

**Q:** Why is the short-run labor demand curve (.hi-purple[D.sub[SR]]) steeper than the short-run labor demand curve (.hi-purple[D.sub[LR]])?

**A:** Firms have fewer alternatives to labor in the short run .mono[-->] less responsive to changes in the wage.
]

---
# Responsiveness

### **Labor demand elasticity**

> A unit-free measure of the responsiveness of the quantity of labor demanded to changes in the wage. 

$$\epsilon_d = \dfrac{\%\Delta E}{\%\Delta w} = \dfrac{(E_2 - E_1) / E_1}{(w_2 - w_1) / w_1} \leq 0$$

--

**Interpretation?** A 1-percent increase in wages .mono[-->] $|\epsilon_d|$-percent decrease in labor demanded.

---
# Responsiveness

### **Labor demand elasticity**

> A unit-free measure of the responsiveness of the quantity of labor demanded to changes in the wage. 

**Q:** What is the labor demand elasticity implied by the data below?

|  Month | Hourly wage | Hours of labor |
|:------:|:-----------:|:--------------:|
| June   | 15          | 120            | 
| July   | 20          | 60             | 

--

**A:** -1.5.

--

$$\epsilon_d = \dfrac{\%\Delta E}{\%\Delta w} = \dfrac{(E_2 - E_1) / E_1}{(w_2 - w_1) / w_1} = \dfrac{(60 - 120) / 120}{(20 - 15) / 15} = \dfrac{-1/2}{1/3} = -1.5$$


---
class: inverse, middle

# Robot tax, featuring Bill Gates

---
class: clear-slide

<iframe 
  src="https://www.youtube.com/embed/nccryZOcrUg" frameborder="0"  allowfullscreen
  allow="
    accelerometer; 
    autoplay; 
    clipboard-write; 
    encrypted-media; 
    gyroscope; 
    picture-in-picture
  "
  style="
    position: fixed;
    top: 0px;
    bottom: 0px;
    right: 0px;
    width: 100%;
    border: none;
    margin: 0;
    padding: 0;
    overflow: hidden;
    z-index: 999999;
    height: 100%;
  ">
  </iframe>

---
# Robot tax, featuring Bill Gates

## **Discussion**

**Q:** How would a robot tax affect... 

- The relative price of robots (capital)?
- The employment of robots (capital) in the long run?
- The employment of labor in the long run?
    - Low-wage labor (*e.g.,* fast food workers, manual laborers, dishwashers, *etc.*)?
    - High-wage labor (*e.g.,* surgeons, engineers, software developers, *etc.*)?
- Tax revenue?


--

**Q:** Should we do it?

---
# Housekeeping

**Problem Set 2** due by Saturday, February 5th by 11:59pm.

**Midterm exam** on Canvas on Monday, February 7th at 2:00pm.

- Extra office hours:
    - Friday from 3:00 to 5:00pm in PLC 522
    - Saturday from 12:00 to 2:00pm on Zoom.
- Study materials (all on Canvas):
    - **Midterm Review Guide**  
    - **Midterm Practice Problems** (w/ key)
    - Key to **Problem Set 1**
    - I will post a key to **Problem Set 2** after the due date


