---
title: "Human Capital"
subtitle: "EC 350: Labor Economics"
author: "[Kyle Raze](https://kyleraze.com)"
date: "Winter 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
      ratio: "16:9"
---

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(ggthemes, readxl, viridis, knitr, ggdag, gapminder, extrafont, tidyverse, magrittr, janitor, kableExtra)
# Define colors
red_pink <- "#e64173"
met_slate <- "#272822" # metropolis font color 
purple <- "#9370DB"
green <- "#007935"
light_green <- "#7DBA97"
orange <- "#FD5F00"
turquoise <- "#44C1C4"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  #dpi = 300,
  #cache = T,
  warning = F,
  message = F
)  
theme_simple <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
theme_market <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.title.x = element_text(hjust = 1, size = 17),
  axis.title.y = element_text(hjust = 1, angle = 0, size = 17),
  # axis.text.x = element_text(size = 12),
  # axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
theme_gif <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)

xaringanExtra::use_scribble()
```

# Human Capital

### **What is it?**

Human capital is set of **acquired skills and experiences** that a worker brings into the labor market.

- **Increases productivity** beyond a worker's innate abilities
- Includes basic literacy and numeracy as well as more-advanced skills
- Non-transferable
- Varies in specificity (*e.g.,* knowing how to code *vs.* knowing how to code in an obscure language)


--

### **Why does it matter?**

Human capital is an important source of **economic growth** and **inequality**.

- Increasing human capital can improve living standards!
- Differences in human capital accumulation generate differences in earnings across workers.

---
# "Typical" age-earnings profiles

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- seq(18, 40, 0.01)

No_College <- function(x) 125*x^2 - 2.4*x^3 - 5000
College <- function(x){ifelse(x >= 22, 185*x^2 + 1000 * x - 3.5*x^3 - 50000, NA)}

base_plot <- ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(18, 40), expand=c(0,0), breaks = c(18, 22.2), labels = c("18", "22")) +
  scale_y_continuous(limits = c(-15000, 75000), label = scales::dollar) +
  theme_bw() +
  theme(text = element_text(family = "Fira Sans", color = met_slate, size = 17),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(margin = margin(r = 10)),
        axis.ticks = element_blank(),
        axis.line = element_line(size = 0.75, color = met_slate),
        axis.line.x = element_blank(),
        axis.text = element_text(size = 18),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(), 
        panel.background = element_rect(fill = "transparent", color = NA),
        plot.background = element_rect(fill = "transparent", color = NA)) +
  labs(x = "Age", y = "Earnings") +
  geom_hline(yintercept = 0, size = 0.75, color = met_slate) +
  stat_function(aes(x), fun = No_College, color = green, size = 0.75) + # no college
  annotate("label", label = "No College", x = 35, y = No_College(35), color = green, family = "Fira Sans", size = 5)

base_plot
```
]
.pull-right[

Earnings increase with experience and eventually decrease with age.

]

---
count: false
# "Typical" age-earnings profiles

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
base_plot +
  stat_function(aes(x), fun = College, color = red_pink, size = 0.75) + # college
  geom_segment(aes(x = 22.2, y = -13000, xend = 22.2, yend = College(22.2)), color = red_pink, size = 0.75, linetype = "solid") +
  geom_segment(aes(x = 18, y = -13000, xend = 22.2, yend = -13000), color = red_pink, size = 0.75) +
  annotate("label", label = "College", x = 35, y = College(35), color = red_pink, family = "Fira Sans", size = 5)
```
]
.pull-right[

Earnings increase with experience and eventually decrease with age.

- **Q:** How does going to college alter this relationship?
]

---
count: false
# "Typical" age-earnings profiles

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
base_plot +
  stat_function(aes(x), fun = College, color = red_pink, size = 0.75) + # college
  geom_segment(aes(x = 22.2, y = -13000, xend = 22.2, yend = College(22.2)), color = red_pink, size = 0.75, linetype = "solid") +
  geom_segment(aes(x = 18, y = -13000, xend = 22.2, yend = -13000), color = red_pink, size = 0.75) +
  annotate("label", label = "College", x = 35, y = College(35), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("text", label = "A", x = 20, y = -6500, color = met_slate, family = "Fira Sans", size = 6) +
  annotate("text", label = "B", x = 20, y = 12500, color = met_slate, family = "Fira Sans", size = 6) +
  annotate("text", label = "C", x = 35, y = 54500, color = met_slate, family = "Fira Sans", size = 6)
```
]
.pull-right[

Earnings increase with experience and eventually decrease with age.

- **Q:** How does going to college alter this relationship?

**Q:** What do areas A, B, and C represent?

]

---
count: false
# "Typical" age-earnings profiles

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
base_plot +
  stat_function(aes(x), fun = College, color = red_pink, size = 0.75) + # college
  geom_segment(aes(x = 22.2, y = -13000, xend = 22.2, yend = College(22.2)), color = red_pink, size = 0.75, linetype = "solid") +
  geom_segment(aes(x = 18, y = -13000, xend = 22.2, yend = -13000), color = red_pink, size = 0.75) +
  annotate("label", label = "College", x = 35, y = College(35), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("text", label = "A", x = 20, y = -6500, color = met_slate, family = "Fira Sans", size = 6) +
  annotate("text", label = "B", x = 20, y = 12500, color = met_slate, family = "Fira Sans", size = 6) +
  annotate("text", label = "C", x = 35, y = 54500, color = met_slate, family = "Fira Sans", size = 6)
```
]
.pull-right[

Earnings increase with experience and eventually decrease with age.

- **Q:** How does going to college alter this relationship?

**Q:** What do areas A, B, and C represent?

- Area A represents the **explicit cost** of college (tuition, books, *etc.*).
- Area B represents the **opportunity cost** of college (forgone earnings).
- Area C represents the monetary **returns to education**.

]

---
# The benefits of education

College is costly!

- Tuition, books, room and board, forgone earnings, stress, *etc.*


--



**Q:** Why did *you* **choose to incur the costs** of going to college? 

--

- To live the life of the mind?
- To increase your earnings potential?
- To expand your social network? 
- To accrue social prestige?
- To set yourself apart?
- To party?
- To find love?


--

While education may have consumption value, we will consider schooling decisions **as investments.**

---
# Education as investment

**Q:** When is it "worth it" to go to college?

- **Benefits?** Going to college causes us to **earn more later in life**.
- **Costs?** Going to college forces us to **forgo earnings now** (and pay tuition, *etc.*).


--

Evaluating this tradeoff requires us to compare dollar amounts spent and received in different time periods.

- To do this, we will use the idea of **present value**, which tells us how much an amount of money received in the future is **worth today**.

---
# Education as investment

## **Present value**

$$\text{PV} = \dfrac{y}{(1 + r)^t}$$

- $y$ is the dollar amount received $t$ periods in the future.
- $r$ is the discount/interest rate.

**The idea?** Getting 100 dollars today is worth more today than getting 100 dollars next year.

- If you got 100 dollars today, you could invest it and end up with $100 \times (1 + r)$ one year from now!

---
# Education as investment

## **Present value**

$$\text{PV} = \dfrac{y}{(1 + r)^t}$$

- $y$ is the dollar amount received $t$ periods in the future.
- $r$ is the discount/interest rate.

**Q:** If the interest rate is 10 percent, what is the present value of receiving 1,000 dollars two years from now?

--

**A:** 826 dollars and 45 cents. $$\begin{align} \text{PV} &= \dfrac{y}{(1 + r)^t} = \dfrac{1000}{(1 + 0.1)^2} = \dfrac{1000}{1.1^2} = \dfrac{1000}{1.21} = 826.45  \end{align}$$

---
# Education as investment

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- seq(18, 40, 0.01)

No_College <- function(x) 30000
College <- function(x){ifelse(x >= 22, 60000, NA)}

base_plot <- ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(18, 40), expand=c(0,0), breaks = c(18, 22.2, 40), labels = c("18", "22", "65")) +
  scale_y_continuous(limits = c(-15000, 75000), breaks = c(-13000, 0, 30000, 60000), labels = c(expression(-H), 0, expression(w[HS]), expression(w[College]))) +
  theme_bw() +
  theme(text = element_text(family = "Fira Sans", color = met_slate, size = 17),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(margin = margin(r = 10)),
        axis.ticks = element_blank(),
        axis.line = element_line(size = 0.75, color = met_slate),
        axis.line.x = element_blank(),
        axis.text = element_text(size = 18),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(), 
        panel.background = element_rect(fill = "transparent", color = NA),
        plot.background = element_rect(fill = "transparent", color = NA)) +
  labs(x = "Age", y = "Earnings") +
  geom_hline(yintercept = 0, size = 0.75, color = met_slate) +
  stat_function(aes(x), fun = No_College, color = green, size = 0.75) + # no college
  annotate("label", label = "No College", x = 35, y = No_College(35), color = green, family = "Fira Sans", size = 5) +
  stat_function(aes(x), fun = College, color = red_pink, size = 0.75) + # college
  geom_segment(aes(x = 22.2, y = -13000, xend = 22.2, yend = College(22.2)), color = red_pink, size = 0.75, linetype = "solid") +
  geom_segment(aes(x = 18, y = -13000, xend = 22.2, yend = -13000), color = red_pink, size = 0.75) +
  annotate("label", label = "College", x = 35, y = College(35), color = red_pink, family = "Fira Sans", size = 5)

base_plot
```
]
.pull-right[
**Q:** When is it "worth it" to go to college?

**A:** Assuming that your objective is to maximize the present value of your **lifetime earnings**, college is worthwhile when $\text{PV}_\text{College} > \text{PV}_\text{HS}$. 

]

---
count: false
# Education as investment

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- seq(18, 40, 0.01)

No_College <- function(x) 30000
College <- function(x){ifelse(x >= 22, 60000, NA)}

base_plot <- ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(18, 40), expand=c(0,0), breaks = c(18, 22.2, 40), labels = c("18", "22", "65")) +
  scale_y_continuous(limits = c(-15000, 75000), breaks = c(-13000, 0, 30000, 60000), labels = c(expression(-H), 0, expression(w[HS]), expression(w[College]))) +
  theme_bw() +
  theme(text = element_text(family = "Fira Sans", color = met_slate, size = 17),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(margin = margin(r = 10)),
        axis.ticks = element_blank(),
        axis.line = element_line(size = 0.75, color = met_slate),
        axis.line.x = element_blank(),
        axis.text = element_text(size = 18),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(), 
        panel.background = element_rect(fill = "transparent", color = NA),
        plot.background = element_rect(fill = "transparent", color = NA)) +
  labs(x = "Age", y = "Earnings") +
  geom_hline(yintercept = 0, size = 0.75, color = met_slate) +
  stat_function(aes(x), fun = No_College, color = green, size = 0.75) + # no college
  annotate("label", label = "No College", x = 35, y = No_College(35), color = green, family = "Fira Sans", size = 5) +
  stat_function(aes(x), fun = College, color = red_pink, size = 0.75) + # college
  geom_segment(aes(x = 22.2, y = -13000, xend = 22.2, yend = College(22.2)), color = red_pink, size = 0.75, linetype = "solid") +
  geom_segment(aes(x = 18, y = -13000, xend = 22.2, yend = -13000), color = red_pink, size = 0.75) +
  annotate("label", label = "College", x = 35, y = College(35), color = red_pink, family = "Fira Sans", size = 5)

base_plot
```
]
.pull-right[
**Q:** When is it "worth it" to go to college?

**A:** Assuming that your objective is to maximize the present value of your **lifetime earnings**, college is worthwhile when $\text{PV}_\text{College} > \text{PV}_\text{HS}$. 

$$\begin{align} \small \text{PV}_\text{HS} &= \small w_\text{HS} + \dfrac{w_\text{HS}}{(1 + r)}+  \dfrac{w_\text{HS}}{(1 + r)^2} + \cdots + \dfrac{w_\text{HS}}{(1 + r)^{46}} \end{align}$$

$$\begin{align} \small \text{PV}_\text{College} = & \small - H - \dfrac{H}{(1 + r)} - \dfrac{H}{(1 + r)^2} - \dfrac{H}{(1 + r)^3} \\  & \small + \small \dfrac{w_\text{College}}{(1 + r)^4} + \dfrac{w_\text{College}}{(1 + r)^5} + \cdots + \dfrac{w_\text{College}}{(1 + r)^{46}} \end{align}$$

]

---
class: clear-slide

**Example:** You are deciding whether to go back to school. 

- You just turned 60, and it would take you 2 years to finish a master's program, which would cost you 10,000 dollars per year.
- You currently earn 80,000 dollars per year. With a master's degree, you could earn 83,000. 
- Regardless of your decision, you are going to retire at 65.

**Q:** If your discount rate is 5 percent, will you choose to go back to school?

---
class: clear-slide

**Example:** You are deciding whether to go back to school. 

- You just turned 60, and it would take you 2 years to finish a master's program, which would cost you 10,000 dollars per year.
- You currently earn 80,000 dollars per year. With a master's degree, you could earn 83,000. 
- Regardless of your decision, you are going to retire at 65.

**Q:** How would the following change your odds of going back to school?

- Your discount rate increases?
- Your post-master's earnings increase?
- Tuition increases?
- You plan to postpone your retirement?


---
# Returns to education

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
wage_school <- function(x) 17500 * log(x + 1)
mrr <- function(x) log(x + 2) / log(x + 1) - 1

plot <- ggplot() +
  stat_function(
    data = tibble(x = 0:20),
    xlim = c(0, 20),
    fun = wage_school,
    aes(x = x),
    color = red_pink,
    size = 1,
  ) +
  scale_x_continuous(limits = c(0, 21), expand = c(0, 0), breaks = seq(0, 20, 4)) +
  scale_y_continuous(expand = c(0, 0), breaks = scales::pretty_breaks(), labels = scales::dollar) +
  labs(x = "Years of education", y = "Earnings") +
  theme_simple +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))
plot
```
]
.pull-right[
## .hi-pink[Wage-schooling locus]

> The amount of money that employers are willing to pay *a particular worker* at every level of schooling. 

]

---
count: false
# Returns to education

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
wage_school <- function(x) 17500 * log(x + 1)
mrr <- function(x) log(x + 2) / log(x + 1) - 1

plot <- ggplot() +
  stat_function(
    data = tibble(x = 0:20),
    xlim = c(0, 20),
    fun = wage_school,
    aes(x = x),
    color = red_pink,
    size = 1,
  ) +
  scale_x_continuous(limits = c(0, 21), expand = c(0, 0), breaks = seq(0, 20, 4)) +
  scale_y_continuous(expand = c(0, 0), breaks = scales::pretty_breaks(), labels = scales::dollar) +
  labs(x = "Years of education", y = "Earnings") +
  theme_simple +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))
plot
```
]
.pull-right[
## .hi-pink[Wage-schooling locus]

> The amount of money that employers are willing to pay *a particular worker* at every level of schooling. 

1. Upward sloping .mono[-->] more school, more money.

2. Slope at a given point .mono[-->] marginal return of an additional year of schooling.

3. Concave .mono[-->] diminishing returns to schooling. 

]

---
# Returns to education

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
mrr <- function(x) 0.85 - 1/30 * x

plot <- ggplot() +
  stat_function(
    data = tibble(x = 0:20),
    xlim = c(1, 20),
    fun = mrr,
    aes(x = x),
    color = red_pink,
    size = 1,
  ) +
  scale_x_continuous(limits = c(0, 21), expand = c(0, 0), breaks = seq(0, 20, 4)) +
  scale_y_continuous(limits = c(0, 1.05), expand = c(0, 0), breaks = scales::pretty_breaks()) +
  labs(x = "Years of education (e)", y = "Rate") +
  theme_simple +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10))) +
  annotate("label", label = "MRR", x = 18, y = mrr(18), color = red_pink, family = "Fira Sans", size = 5)

plot
```
]
.pull-right[
## .hi-pink[Marginal rate of return]

> The percentage increase in earnings from an additional year of schooling: $$\text{MRR} = \dfrac{\%\Delta w}{\Delta e}$$

]

---
# Schooling decisions

## **The stopping rule**

**Q:** How does a worker choose the optimal.super[.hi-pink[<span>&#8224;</span>]] amount of schooling?

.footnote[.super[.hi-pink[<span>&#8224;</span>]] "Optimal" in the sense of maximizing the present value of lifetime earnings.]

--

**A:** A worker chooses the optimal amount of schooling $e^*$ where the marginal rate of return equals the discount rate: $$\text{MRR} = r$$

--

- If $\text{MRR} > r$, then schooling education would increase the present value of lifetime earnings.
- If $\text{MRR} < r$, then the worker has "gone too far"&mdash;she could had a higher present value of lifetime earnings if she had completed less schooling.


---
# Schooling decisions

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:20
mrr <- function(x) 1 - 1/20 * x
discount <- 0.55
optimal_e <- uniroot(function(x) mrr(x) - discount, range(x))$root

ggplot() +
  stat_function(
    data = tibble(x = 0:20),
    xlim = c(1, 20),
    fun = mrr,
    aes(x = x),
    color = red_pink,
    size = 1,
  ) +
  annotate("label", label = "MRR", x = 18, y = mrr(18), color = red_pink, family = "Fira Sans", size = 5) +
  geom_hline(yintercept = discount, color = met_slate, size = 1) +
  geom_point(aes(x = optimal_e, y = discount), color = met_slate, size = 2) +
  geom_segment(aes(x = optimal_e, y = 0, xend = optimal_e, yend = discount), linetype  = "dashed", color = met_slate) + 
  scale_x_continuous(limits = c(0, 21), expand = c(0, 0), breaks = c(optimal_e), labels = c(expression(e^"*"))) +
  scale_y_continuous(limits = c(0, 1.05), expand = c(0, 0), breaks = c(discount), labels = c(expression(r))) +
  labs(x = "Years of education (e)", y = "Discount rate") +
  theme_simple +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))
```
]
.pull-right[
## **The stopping rule**

A worker chooses the optimal amount of schooling where the marginal rate of return intersects the discount rate.
]

---
# Comparing schooling decisions

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:20
mrr <- function(x) 1 - 1/20 * x
discount_1 <- 0.75
discount_2 <- 0.25
optimal_e_1 <- uniroot(function(x) mrr(x) - discount_1, range(x))$root
optimal_e_2 <- uniroot(function(x) mrr(x) - discount_2, range(x))$root

plot <- ggplot() +
  stat_function(
    data = tibble(x = 0:20),
    xlim = c(1, 20),
    fun = mrr,
    aes(x = x),
    color = red_pink,
    size = 1,
  ) +
  annotate("label", label = expression(MRR[1]==MRR[2]), x = 18, y = mrr(18), color = red_pink, family = "Fira Sans", size = 5) +
  geom_hline(yintercept = discount_1, color = met_slate, size = 1) +
  geom_hline(yintercept = discount_2, color = met_slate, size = 1) +
  geom_point(aes(x = optimal_e_1, y = discount_1), color = met_slate, size = 2) +
  geom_segment(aes(x = optimal_e_1, y = 0, xend = optimal_e_1, yend = discount_1), linetype  = "dashed", color = met_slate) + 
  geom_point(aes(x = optimal_e_2, y = discount_2), color = met_slate, size = 2) +
  geom_segment(aes(x = optimal_e_2, y = 0, xend = optimal_e_2, yend = discount_2), linetype  = "dashed", color = met_slate) + 
  scale_x_continuous(limits = c(0, 21), expand = c(0, 0), breaks = c(optimal_e_1, optimal_e_2), labels = c(expression(e[1]), expression(e[2]))) +
  scale_y_continuous(limits = c(0, 1.05), expand = c(0, 0), breaks = c(discount_2, discount_1), labels = c(expression(r[2]), expression(r[1]))) +
  labs(x = "Years of education (e)", y = "Discount rate") +
  theme_simple +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))

plot
```
]
.pull-right[

## **Differences in discount rates**

Higher discount rate .mono[-->] less access to credit or stronger preferences toward immediate payoffs.

- Given two individuals with the same ability, the person with a **higher discount rate** will complete **fewer years** of schooling.
- In either case, the person with the higher discount rate will earn less money.




]

---
count: false
# Comparing schooling decisions

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
plot
```
]
.pull-right[

## **Differences in discount rates**

Higher discount rate .mono[-->] less access to credit or stronger preferences toward immediate payoffs.

- Given two individuals with the same ability, the person with a **higher discount rate** will complete **fewer years** of schooling.
- In either case, the person with the higher discount rate will earn less money.

**Implications for policy?** Expanding educational opportunities to person with the higher discount rate will close the earnings gap!

]

---
# Comparing schooling decisions

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:20
mrr_1 <- function(x) 0.75 - 1/20 * x
mrr_2 <- function(x) 1.25 - 1/20 * x
discount_1 <- 0.5
discount_2 <- 0.5
optimal_e_1 <- uniroot(function(x) mrr_1(x) - discount_1, range(x))$root
optimal_e_2 <- uniroot(function(x) mrr_2(x) - discount_2, range(x))$root

plot <- ggplot() +
  stat_function(
    data = tibble(x = 0:20),
    xlim = c(1, 20),
    fun = mrr_1,
    aes(x = x),
    color = red_pink,
    size = 1,
  ) +
  annotate("label", label = expression(MRR[1]), x = 9, y = mrr_1(9), color = red_pink, family = "Fira Sans", size = 5) +
  
  stat_function(
    data = tibble(x = 0:20),
    xlim = c(1, 20),
    fun = mrr_2,
    aes(x = x),
    color = purple,
    size = 1,
  ) +
  annotate("label", label = expression(MRR[2]), x = 19, y = mrr_2(19), color = purple, family = "Fira Sans", size = 5) +
  
  geom_hline(yintercept = discount_1, color = met_slate, size = 1) +
  geom_point(aes(x = optimal_e_1, y = discount_1), color = met_slate, size = 2) +
  geom_segment(aes(x = optimal_e_1, y = 0, xend = optimal_e_1, yend = discount_1), linetype  = "dashed", color = met_slate) + 
  geom_point(aes(x = optimal_e_2, y = discount_2), color = met_slate, size = 2) +
  geom_segment(aes(x = optimal_e_2, y = 0, xend = optimal_e_2, yend = discount_2), linetype  = "dashed", color = met_slate) + 
  scale_x_continuous(limits = c(0, 21), expand = c(0, 0), breaks = c(optimal_e_1, optimal_e_2), labels = c(expression(e[1]), expression(e[2]))) +
  scale_y_continuous(limits = c(0, 1.05), expand = c(0, 0), breaks = c(discount_1), labels = c(expression(r))) +
  labs(x = "Years of education (e)", y = "Discount rate") +
  theme_simple +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))

plot
```
]
.pull-right[

## **Differences in ability**

A .purple[higher-ability individual] "gets more" out of the same amount of schooling than a .pink[lower-ability individual] .mono[-->] higher marginal rate of return.

- Given the same discount rate, then the .hi-purple[higher-ability individual] will complete **more schooling** and earn **more money**.
]

---
count: false
# Comparing schooling decisions

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
plot
```
]
.pull-right[

## **Differences in ability**

A .purple[higher-ability individual] "gets more" out of the same amount of schooling than a .pink[lower-ability individual] .mono[-->] higher marginal rate of return.

- Given the same discount rate, then the .hi-purple[higher-ability individual] will complete **more schooling** and earn **more money**.


**Implications for policy?** Closing the schooling gap won't close the earnings gap!

]

---
count: false
# Comparing schooling decisions

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
plot
```
]
.pull-right[

## **Differences in ability**

A .purple[higher-ability individual] "gets more" out of the same amount of schooling than a .pink[lower-ability individual] .mono[-->] higher marginal rate of return.

- Given the same discount rate, then the .hi-purple[higher-ability individual] will complete **more schooling** and earn **more money**.


**Implications for policy?** Closing the schooling gap won't close the earnings gap!

**Implications for data analysis?**

]

---
# Arteaga (2018)

## **Discussion**

**Q.sub[1]:** What is the research question? Why does it matter?

**Q.sub[2]:** What is the research design? What are the comparison groups?

**Q.sub[3]:** What are the main results? What story do they convey?

---
# Housekeeping

Problem Set 3 is due by **~~Friday, February 25th~~** **Monday, February 28th at 11:59pm**.
