---
title: "Discrimination"
subtitle: "EC 350: Labor Economics"
author: "[Kyle Raze](https://kyleraze.com)"
date: "Winter 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
      ratio: "16:9"
---

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(ggthemes, readxl, viridis, knitr, ggdag, gapminder, extrafont, tidyverse, magrittr, janitor, kableExtra)
# Define colors
red_pink <- "#e64173"
met_slate <- "#272822" # metropolis font color 
purple <- "#9370DB"
green <- "#007935"
light_green <- "#7DBA97"
orange <- "#FD5F00"
turquoise <- "#44C1C4"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  #dpi = 300,
  #cache = T,
  warning = F,
  message = F
)  
theme_simple <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  # strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
theme_market <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.title.x = element_text(hjust = 1, size = 17),
  axis.title.y = element_text(hjust = 1, angle = 0, size = 17),
  # axis.text.x = element_text(size = 12),
  # axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
theme_gif <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)

xaringanExtra::use_scribble()
```

# Discrimination

### **What is it?**

Labor market discrimination occurs when two or more **equally productive individuals** are **treated differently** on the basis of some other characteristic.

- **Examples?** Age, race, gender, *etc.*


--

### **Why should we care?**

Discrimination is...

- Morally objectionable.
- Illegal.
- A cause of inequality.
- Common.




---
# Understanding group differences

**Q:** What do we mean when we say ".hi-pink[Group A] **earns less than** .hi-green[Group B]**, on average**?"

---
count: false
# Understanding group differences

**Q:** What do we mean when we say ".hi-pink[Group A] **earns less than** .hi-green[Group B]**, on average**?"

```{r, echo = F, dev = "svg", fig.height = 5.25}
diff <- 1
df <- tibble(x = seq(-4, 4, 0.01), y = dnorm(x)) %>%
  rbind(., tibble(x = seq(4, -4, -0.01), y = 0)) %>% 
  mutate(x2 = x + diff)
ggplot() +
  scale_x_continuous(limits = c(-4, 5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 0.45), expand = c(0, 0)) +
  geom_polygon(data = df, aes(x, y), fill = red_pink, alpha = 0.5) +
  # geom_polygon(data = df, aes(x = x2, y), fill = green, alpha = 0.5) +
  # geom_vline(xintercept = diff, size = 1, color = green) +
  geom_vline(xintercept = 0, size = 1, color = red_pink) +
  theme_simple +
  theme(axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text.x = element_blank(), axis.title = element_blank())
```

---
count: false
# Understanding group differences

**Q:** What do we mean when we say ".hi-pink[Group A] **earns less than** .hi-green[Group B]**, on average**?"

```{r, echo = F, dev = "svg", fig.height = 5.25}
diff <- 1
df <- tibble(x = seq(-4, 4, 0.01), y = dnorm(x)) %>%
  rbind(., tibble(x = seq(4, -4, -0.01), y = 0)) %>% 
  mutate(x2 = x + diff)
ggplot() +
  scale_x_continuous(limits = c(-4, 5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 0.45), expand = c(0, 0)) +
  geom_polygon(data = df, aes(x, y), fill = red_pink, alpha = 0.5) +
  geom_polygon(data = df, aes(x = x2, y), fill = green, alpha = 0.5) +
  geom_vline(xintercept = diff, size = 1, color = green) +
  geom_vline(xintercept = 0, size = 1, color = red_pink) +
  theme_simple +
  theme(axis.line.y = element_blank(), axis.text.y = element_blank(), axis.text.x = element_blank(), axis.title = element_blank())
```


---
# Wage gaps

## **Data**

**The source?** Current Population Survey (via [IPUMS](https://cps.ipums.org/cps/)).

**The sample?** Nationally representative sample of employed working-age adults aged 16.mono[+].

- We will further restrict the sample to those who self-identify as Asian, Black, Hispanic (of any race), or White (non-Hispanic).
- We will first analyze the 2021 data through April (23,556 survey respondents), then incorporate data going back to 2010 (999,313 survey respondents).

**The outcome?** Hourly wages.

---
# Wage gaps

## **Approach**

Run regressions of the form $$\log(\text{Wage}_i) = \alpha + \beta~\text{Group}_i + X^{'} \Phi + \varepsilon_i$$

- $\text{Group}_i$ is an indicator variable denoting race or gender.
- $\beta \times 100$ is the wage gap, interpreted as a percentage difference.
    - Relative to a "reference group."
- $X^{'}$ is a vector of control variables.


---
class: clear-slide

.center[**Racial wage gaps among men (2021)**]
```{R, echo = F, escape = F}
tab <- tibble(
  v1 = c("Asian", "", "Black", "", "Hispanic", "", 
         "Education",
         "Experience", 
         "Industry",
         "Occupation",
         "State"),
  v2 = c(0.025, "(0.028)", -0.191, "(0.016)", -0.122, "(0.011)", "", "", "", "", ""),
  v3 = c(-0.022, "(0.026)", -0.178, "(0.015)", -0.051, "(0.011)", "$\\checkmark$", "", "", "", ""),
  v4 = c(-0.022, "(0.025)", -0.175, "(0.014)", -0.034, "(0.011)", "$\\checkmark$", "$\\checkmark$", "", "", ""),
  v5 = c(-0.002, "(0.025)", -0.145, "(0.014)", -0.032, "(0.010)", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "", ""),
  v6 = c(0.004, "(0.022)", -0.115, "(0.014)", -0.017, "(0.010)", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", ""),
  v7 = c(-0.027, "(0.022)", -0.095, "(0.016)", -0.011, "(0.015)", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$"),
) %>% kable(
  escape = FALSE,
  col.names = c("", "(1)", "(2)", "(3)", "(4)", "(5)", "(6)"),
  align = c("l", rep("c", 6)),
  table.attr = "style='width:100%;'"
) %>%
row_spec(0, color = "#708090") %>% 
row_spec(1:11, color = met_slate) %>%
row_spec(seq(2,6,2), color = "#c2bebe") %>%
row_spec(1:11, extra_css = "line-height: 110%;") %>%
column_spec(1, color = met_slate, italic = T)

tab
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among men (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(2, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among men (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(3, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among men (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(4, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among men (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(5, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among men (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(6, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among men (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(7, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
class: clear-slide

.center[**Racial wage gaps among men (2010-2021)**]
```{r, echo = F, dev = "svg", fig.height = 5.5}
df <- readRDS("annual_race_wage_gaps.rds") %>%
  mutate(
    gender = case_when(female == 0 ~ "Men", T ~ "Women"),
    term = case_when(
      term == "asian" ~ "Asian-White",
      term == "black" ~ "Black-White",
      term == "hispanic" ~ "Hispanic-White"
    )
  )


df %>% 
  filter(gender == "Men") %>% 
  ggplot(aes(
      x = year,
      y = estimate,
      color = as.factor(term),
      shape = as.factor(analysis)
    )) +
    geom_hline(yintercept = 0) +
    geom_linerange(
      aes(ymin = conf.low, ymax = conf.high),
      size = 1.5,
      alpha = 0.5,
      position = position_dodge2(preserve = "single", width = 0.75)
    ) +
    geom_point(
      size = 1.5,
      fill = "white",
      stroke = 2,
      position = position_dodge2(preserve = "single", width = 0.75)
    ) +
  facet_grid(~ term, scales = "free_y") +
  scale_y_continuous(limits = c(-0.25, 0.15),
                       breaks = scales::pretty_breaks(n = 6)) +
  scale_shape_manual(values = c(21, 24)) +
  scale_color_manual(values = c(red_pink, purple, green)) +
  guides(color = F) +
  labs(y = "Gap estimate", shape = "Gap") +
  theme_simple +
  theme(
    strip.background = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Vertical bars outline 95% confidence intervals. Reference category .mono[=] White men.]

---
class: clear-slide

.center[**Racial wage gaps among women (2021)**]
```{R, echo = F, escape = F}
tab <- tibble(
  v1 = c("Asian", "", "Black", "", "Hispanic", "", 
         "Education",
         "Experience", 
         "Industry",
         "Occupation",
         "State"),
  v2 = c(0.065, "(0.025)", -0.102, "(0.014)", -0.123, "(0.012)", "", "", "", "", ""),
  v3 = c(0.019, "(0.022)", -0.063, "(0.012)", -0.013, "(0.011)", "$\\checkmark$", "", "", "", ""),
  v4 = c(0.011, "(0.022)", -0.057, "(0.012)", 0.004, "(0.011)", "$\\checkmark$", "$\\checkmark$", "", "", ""),
  v5 = c(0.019, "(0.021)", -0.069, "(0.012)", 0.013, "(0.011)", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "", ""),
  v6 = c(0.039, "(0.020)", -0.040, "(0.012)", 0.034, "(0.010)", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", ""),
  v7 = c(-0.024, "(0.021)", -0.023, "(0.012)", 0.005, "(0.011)", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$"),
) %>% kable(
  escape = FALSE,
  col.names = c("", "(1)", "(2)", "(3)", "(4)", "(5)", "(6)"),
  align = c("l", rep("c", 6)),
  table.attr = "style='width:100%;'"
) %>%
row_spec(0, color = "#708090") %>% 
row_spec(1:11, color = met_slate) %>%
row_spec(seq(2,6,2), color = "#c2bebe") %>%
row_spec(1:11, extra_css = "line-height: 110%;") %>%
column_spec(1, color = met_slate, italic = T)

tab
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White women.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among women (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(2, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White women.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among women (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(3, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White women.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among women (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(4, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White women.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among women (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(5, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White women.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among women (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(6, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White women.]

---
count: false
class: clear-slide

.center[**Racial wage gaps among women (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(7, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White women.]

---
class: clear-slide

.center[**Racial wage gaps among women (2010-2021)**]
```{r, echo = F, dev = "svg", fig.height = 5.5}
df <- readRDS("annual_race_wage_gaps.rds") %>%
  mutate(
    gender = case_when(female == 0 ~ "Men", T ~ "Women"),
    term = case_when(
      term == "asian" ~ "Asian-White",
      term == "black" ~ "Black-White",
      term == "hispanic" ~ "Hispanic-White"
    )
  )


df %>% 
  filter(gender == "Women") %>% 
  ggplot(aes(
      x = year,
      y = estimate,
      color = as.factor(term),
      shape = as.factor(analysis)
    )) +
    geom_hline(yintercept = 0) +
    geom_linerange(
      aes(ymin = conf.low, ymax = conf.high),
      size = 1.5,
      alpha = 0.5,
      position = position_dodge2(preserve = "single", width = 0.75)
    ) +
    geom_point(
      size = 1.5,
      fill = "white",
      stroke = 2,
      position = position_dodge2(preserve = "single", width = 0.75)
    ) +
  facet_grid(~ term, scales = "free_y") +
  scale_y_continuous(limits = c(-0.25, 0.15),
                       breaks = scales::pretty_breaks(n = 6)) +
  scale_shape_manual(values = c(21, 24)) +
  scale_color_manual(values = c(red_pink, purple, green)) +
  guides(color = F) +
  labs(y = "Gap estimate", shape = "Gap") +
  theme_simple +
  theme(
    strip.background = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Vertical bars outline 95% confidence intervals. Reference category .mono[=] White women.]


---
class: clear-slide

.center[**Gender wage gap among Whites (2021)**]
```{R, echo = F, escape = F}
tab <- tibble(
  v1 = c("Female", "", 
         "Education",
         "Experience", 
         "Industry",
         "Occupation",
         "State"),
  v2 = c(-0.114, "(0.009)", "", "", "", "", ""),
  v3 = c(-0.159, "(0.009)", "$\\checkmark$", "", "", "", ""),
  v4 = c(-0.167, "(0.008)", "$\\checkmark$", "$\\checkmark$", "", "", ""),
  v5 = c(-0.121, "(0.009)", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "", ""),
  v6 = c(-0.090, "(0.009)", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", ""),
  v7 = c(-0.086, "(0.009)", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$")
) %>% kable(
  escape = FALSE,
  col.names = c("", "(1)", "(2)", "(3)", "(4)", "(5)", "(6)"),
  align = c("l", rep("c", 6)),
  table.attr = "style='width:100%;'"
) %>%
row_spec(0, color = "#708090") %>% 
row_spec(1:7, color = met_slate) %>%
row_spec(2, color = "#c2bebe") %>%
row_spec(1:7, extra_css = "line-height: 110%;") %>%
column_spec(1, color = met_slate, italic = T)

tab
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Gender wage gap among Whites (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(2, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Gender wage gap among Whites (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(3, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Gender wage gap among Whites (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(4, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Gender wage gap among Whites (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(5, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Gender wage gap among Whites (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(6, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]

---
count: false
class: clear-slide

.center[**Gender wage gap among Whites (2021)**]
```{R, echo = F, escape = F}
tab %>% column_spec(7, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Standard errors in parentheses. Reference category .mono[=] White men.]


---
class: clear-slide

.center[**Gender wage gaps by race (2010-2021)**]
```{r, echo = F, dev = "svg", fig.height = 5.5}
df <- readRDS("annual_gender_wage_gaps.rds")

df %>% 
  ggplot(aes(
      x = year,
      y = estimate,
      color = as.factor(race),
      shape = as.factor(analysis)
    )) +
    geom_hline(yintercept = 0) +
    geom_linerange(
      aes(ymin = conf.low, ymax = conf.high),
      size = 1.5,
      alpha = 0.5,
      position = position_dodge2(preserve = "single", width = 1)
    ) +
    geom_point(
      size = 1.5,
      fill = "white",
      stroke = 2,
      position = position_dodge2(preserve = "single", width = 1)
    ) +
  facet_grid(~ race, scales = "free_y") +
  scale_y_continuous(limits = c(-0.2, 0.075),
                       breaks = scales::pretty_breaks(n = 6)) +
  scale_shape_manual(values = c(21, 24)) +
  scale_color_manual(values = c(red_pink, purple, green, met_slate)) +
  guides(color = F) +
  labs(y = "Gap estimate", shape = "Gap") +
  theme_simple +
  theme(
    strip.background = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```
.footnote[*Notes:* Outcome variable .mono[=] log(hourly wage). Vertical bars outline 95% confidence intervals. Reference category .mono[=] Men.]

---
# Wage gaps

## **Discussion**

**Q:** Do the adjusted gaps provide causal evidence of gender or race discrimination? Why or why not?

- Do the adjusted gaps control for enough factors to make an *all-else-equal* claim?
- Do the adjusted gaps adjust for too much?

---
# Collider bias

## **Example: Gender discrimination**

.pull-left[
**Variables**
- .hi[F] .mono[=] Gender
- .hi[D] .mono[=] Discrimination
- .hi[Y] .mono[=] Wages
- .hi[O] .mono[=] Occupation
- .hi[U] .mono[=] Unobserved ability

**Causal paths**
- .hi[D] .mono[-->] .hi[Y] (path of interest)
- .hi[D] .mono[-->] .hi[O] .mono[-->] .hi[Y] (mediated path)
- .hi[D] .mono[-->] .hi[O] .mono[<--] .hi[A] .mono[-->] .hi[Y]
]
.pull-right[
```{r, dev = "svg", echo = F, fig.width = 5.5, fig.height = 5}
set.seed(1229)
dag <-
  dagify(Y ~ D + O + A,
         D ~ `F`,
         O ~ D + A,
         exposure = "D",
         outcome = "Y") %>%
  ggplot(aes(
    x = x,
    y = y,
    xend = xend,
    yend = yend
  )) +
  geom_dag_point(color = red_pink) +
  geom_dag_edges_link(arrow = grid::arrow(length = grid::unit(10, "pt"), type = "closed")) +
  geom_dag_text(family = "Fira Sans", size = 10) +
  theme_dag()
dag
```
]

---
# Collider bias

## **Example: Gender discrimination**

.pull-left[
**Simulation**
```{R, eval = F}
simulated_data <- tibble(
  female = ifelse(runif(10000) >= 0.5, 1, 0),
  ability = rnorm(10000),
  discrimination = female,
  occupation = 1 + 2 * ability + 0 * female - 2 * discrimination + rnorm(10000),
  wage = 1 - 1 * discrimination + 1 * occupation + 2 * ability + rnorm(10000)
)

lm_1 <- lm(wage ~ female, 
           data = simulated_data)

lm_2 <- lm(wage ~ female + occupation, 
           data = simulated_data)

lm_3 <- lm(wage ~ female + occupation + ability, 
           data = simulated_data)
```

]
.pull-right[
```{r, dev = "svg", echo = F, fig.width = 5.5, fig.height = 5}
dag
```
]

---
count: false
# Collider bias

## **Example: Gender discrimination**

.pull-left[
**Simulation**
```{R, eval = F}
simulated_data <- tibble(
  female = ifelse(runif(10000) >= 0.5, 1, 0), #<<
  ability = rnorm(10000),
  discrimination = female,
  occupation = 1 + 2 * ability + 0 * female - 2 * discrimination + rnorm(10000),
  wage = 1 - 1 * discrimination + 1 * occupation + 2 * ability + rnorm(10000)
)

lm_1 <- lm(wage ~ female, 
           data = simulated_data)

lm_2 <- lm(wage ~ female + occupation, 
           data = simulated_data)

lm_3 <- lm(wage ~ female + occupation + ability, 
           data = simulated_data)
```

]
.pull-right[
```{r, dev = "svg", echo = F, fig.width = 5.5, fig.height = 5}
dag
```
]

---
count: false
# Collider bias

## **Example: Gender discrimination**

.pull-left[
**Simulation**
```{R, eval = F}
simulated_data <- tibble(
  female = ifelse(runif(10000) >= 0.5, 1, 0),
  ability = rnorm(10000), #<<
  discrimination = female,
  occupation = 1 + 2 * ability + 0 * female - 2 * discrimination + rnorm(10000),
  wage = 1 - 1 * discrimination + 1 * occupation + 2 * ability + rnorm(10000)
)

lm_1 <- lm(wage ~ female, 
           data = simulated_data)

lm_2 <- lm(wage ~ female + occupation, 
           data = simulated_data)

lm_3 <- lm(wage ~ female + occupation + ability, 
           data = simulated_data)
```

]
.pull-right[
```{r, dev = "svg", echo = F, fig.width = 5.5, fig.height = 5}
dag
```
]

---
count: false
# Collider bias

## **Example: Gender discrimination**

.pull-left[
**Simulation**
```{R, eval = F}
simulated_data <- tibble(
  female = ifelse(runif(10000) >= 0.5, 1, 0),
  ability = rnorm(10000),
  discrimination = female, #<<
  occupation = 1 + 2 * ability + 0 * female - 2 * discrimination + rnorm(10000),
  wage = 1 - 1 * discrimination + 1 * occupation + 2 * ability + rnorm(10000)
)

lm_1 <- lm(wage ~ female, 
           data = simulated_data)

lm_2 <- lm(wage ~ female + occupation, 
           data = simulated_data)

lm_3 <- lm(wage ~ female + occupation + ability, 
           data = simulated_data)
```

]
.pull-right[
```{r, dev = "svg", echo = F, fig.width = 5.5, fig.height = 5}
dag
```
]

---
count: false
# Collider bias

## **Example: Gender discrimination**

.pull-left[
**Simulation**
```{R, eval = F}
simulated_data <- tibble(
  female = ifelse(runif(10000) >= 0.5, 1, 0),
  ability = rnorm(10000),
  discrimination = female,
  occupation = 1 + 2 * ability + 0 * female - 2 * discrimination + rnorm(10000), #<<
  wage = 1 - 1 * discrimination + 1 * occupation + 2 * ability + rnorm(10000)
)

lm_1 <- lm(wage ~ female, 
           data = simulated_data)

lm_2 <- lm(wage ~ female + occupation, 
           data = simulated_data)

lm_3 <- lm(wage ~ female + occupation + ability, 
           data = simulated_data)
```

]
.pull-right[
```{r, dev = "svg", echo = F, fig.width = 5.5, fig.height = 5}
dag
```
]

---
count: false
# Collider bias

## **Example: Gender discrimination**

.pull-left[
**Simulation**
```{R, eval = F}
simulated_data <- tibble(
  female = ifelse(runif(10000) >= 0.5, 1, 0),
  ability = rnorm(10000),
  discrimination = female,
  occupation = 1 + 2 * ability + 0 * female - 2 * discrimination + rnorm(10000),
  wage = 1 - 1 * discrimination + 1 * occupation + 2 * ability + rnorm(10000) #<<
)

lm_1 <- lm(wage ~ female, 
           data = simulated_data)

lm_2 <- lm(wage ~ female + occupation, 
           data = simulated_data)

lm_3 <- lm(wage ~ female + occupation + ability, 
           data = simulated_data)
```

]
.pull-right[
```{r, dev = "svg", echo = F, fig.width = 5.5, fig.height = 5}
dag
```
]

---
# Collider bias

## **Example: Gender discrimination**

.pull-left[
**Simulation**
```{R, eval = F}
simulated_data <- tibble(
  female = ifelse(runif(10000) >= 0.5, 1, 0),
  ability = rnorm(10000),
  discrimination = female,
  occupation = 1 + 2 * ability + 0 * female - 2 * discrimination + rnorm(10000),
  wage = 1 - 1 * discrimination + 1 * occupation + 2 * ability + rnorm(10000)
)

lm_1 <- lm(wage ~ female, 
           data = simulated_data)

lm_2 <- lm(wage ~ female + occupation, 
           data = simulated_data)

lm_3 <- lm(wage ~ female + occupation + ability, 
           data = simulated_data)
```

]
.pull-right[
**Results**
```{R, echo = F, escape = F}
p_load(broom)

set.seed(6399)

simulated_data <- tibble(
  female = ifelse(runif(10000) >= 0.5, 1, 0),
  ability = rnorm(10000),
  discrimination = female,
  occupation = 1 + 2 * ability + 0 * female - 2 * discrimination + rnorm(10000),
  wage = 1 - 1 * discrimination + 1 * occupation + 2 * ability + rnorm(10000)
)

lm_1 <- lm(wage ~ female, data = simulated_data) %>%
  tidy() %>% 
  mutate(estimate = round(estimate, 2), std.error = round(std.error, 2))

lm_2 <- lm(wage ~ female + occupation, data = simulated_data) %>%
  tidy() %>% 
  mutate(estimate = round(estimate, 2), std.error = round(std.error, 2))

lm_3 <- lm(wage ~ female + occupation + ability, data = simulated_data) %>%
  tidy() %>% 
  mutate(estimate = round(estimate, 2), std.error = round(std.error, 2))

tab <- data.frame(
  v1 = c("Female", "", 
         "Occupation", "",
         "Ability", "",
         "Intercept", ""),
  v2 = c(-3.05, "(0.09)", "", "", "", "", 1.95, "(0.06)"),
  v3 = c(0.58, "(0.03)", 1.80, "(0.01)", "", "", 0.20, "(0.02)"),
  v4 = c(-1.01, "(0.03)", 1.00, "(0.01)", 2.00, "(0.02)", 1.01, "(0.02)")
) %>% kable(
  escape = FALSE,
  col.names = c("", "(1)", "(2)", "(3)"),
  align = c("l", rep("c", 3)),
  table.attr = "style='width:100%;'"
) %>%
row_spec(0, color = "#708090") %>% 
row_spec(1:8, color = met_slate) %>%
row_spec(c(seq(2, 8, 2)), color = "#c2bebe") %>%
row_spec(1:8, extra_css = "line-height: 110%;") %>%
column_spec(1, color = met_slate, italic = T)

tab
```
.footnote[*Notes:* Outcome variable .mono[=] wage. Standard errors in parentheses. Reference category .mono[=] Men.]
]

---
count: false
# Collider bias

## **Example: Gender discrimination**

.pull-left[
**Simulation**
```{R, eval = F}
simulated_data <- tibble(
  female = ifelse(runif(10000) >= 0.5, 1, 0),
  ability = rnorm(10000),
  discrimination = female,
  occupation = 1 + 2 * ability + 0 * female - 2 * discrimination + rnorm(10000),
  wage = 1 - 1 * discrimination + 1 * occupation + 2 * ability + rnorm(10000)
)

lm_1 <- lm(wage ~ female, 
           data = simulated_data)

lm_2 <- lm(wage ~ female + occupation, 
           data = simulated_data)

lm_3 <- lm(wage ~ female + occupation + ability, 
           data = simulated_data)
```

]
.pull-right[
**Results**
```{R, echo = F, escape = F}
tab %>% column_spec(2, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] wage. Standard errors in parentheses. Reference category .mono[=] Men.]
]

---
count: false
# Collider bias

## **Example: Gender discrimination**

.pull-left[
**Simulation**
```{R, eval = F}
simulated_data <- tibble(
  female = ifelse(runif(10000) >= 0.5, 1, 0),
  ability = rnorm(10000),
  discrimination = female,
  occupation = 1 + 2 * ability + 0 * female - 2 * discrimination + rnorm(10000),
  wage = 1 - 1 * discrimination + 1 * occupation + 2 * ability + rnorm(10000)
)

lm_1 <- lm(wage ~ female, 
           data = simulated_data)

lm_2 <- lm(wage ~ female + occupation, 
           data = simulated_data)

lm_3 <- lm(wage ~ female + occupation + ability, 
           data = simulated_data)
```

]
.pull-right[
**Results**
```{R, echo = F, escape = F}
tab %>% column_spec(3, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] wage. Standard errors in parentheses. Reference category .mono[=] Men.]
]

---
count: false
# Collider bias

## **Example: Gender discrimination**

.pull-left[
**Simulation**
```{R, eval = F}
simulated_data <- tibble(
  female = ifelse(runif(10000) >= 0.5, 1, 0),
  ability = rnorm(10000),
  discrimination = female,
  occupation = 1 + 2 * ability + 0 * female - 2 * discrimination + rnorm(10000),
  wage = 1 - 1 * discrimination + 1 * occupation + 2 * ability + rnorm(10000)
)

lm_1 <- lm(wage ~ female, 
           data = simulated_data)

lm_2 <- lm(wage ~ female + occupation, 
           data = simulated_data)

lm_3 <- lm(wage ~ female + occupation + ability, 
           data = simulated_data)
```

]
.pull-right[
**Results**
```{R, echo = F, escape = F}
tab %>% column_spec(4, bold = T)
```
.footnote[*Notes:* Outcome variable .mono[=] wage. Standard errors in parentheses. Reference category .mono[=] Men.]
]

---
# Bertrand and Mullainathan (2004)

## **Discussion**

**Q.sub[1]:** How does the study measure discrimination in the labor market? 

**Q.sub[2]:** What are the advantages of the research design?

**Q.sub[3]:** What are the weaknesses of the study?

**Q.sub[4]:** What are the main findings? 

**Q.sub[5]:** What does the study tell us about employers?

**Q.sub[6]:** What did *you* find most interesting and/or depressing?

---
# Housekeeping

**Assigned reading for Wednesday:** [Ban the Box, Criminal Records, and Racial Discrimination: A Field Experiment](https://academic.oup.com/qje/article-abstract/133/1/191/4060073) by Amanda Agan and Sonja Starr (2017).

- **Reading Quiz 10** is due by **Monday, March 7th at 12pm (noon)**.
- The quiz instructions will include a reading guide.
