---
title: "Data and Causation"
subtitle: "EC 350: Labor Economics"
author: "[Kyle Raze](https://kyleraze.com)"
date: "Spring 2021"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
      ratio: "16:9"
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE, knitr.table.format = "html")
library(pacman)
p_load(ggthemes, viridis, knitr, emoGG, kableExtra, gridExtra, extrafont, webshot, fontawesome, tidyverse, magrittr, janitor, wooldridge)
# Define colors
red_pink <- "#e64173"
met_slate <- "#272822" # metropolis font color 
purple <- "#9370DB"
green <- "#007935"
light_green <- "#7DBA97"
# Notes directory
dir_slides <- "~/GitHub/Intro_Labor_S21/Lectures/03-Data_Causation/"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  #dpi = 300,
  #cache = T,
  warning = F,
  message = F
)  
theme_simple <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
wrapper <- function(x, ...) paste(strwrap(x, ...), collapse = "\n")
```

# Prologue

---
# Data and Causation

1. The rise of empirical evidence
2. Making *other-things-equal* comparisons
3. What do "data" look like?
4. Causal identification
    - Average treatment effects
    - Selection bias
5. Randomized control trials
6. *Thinking Fast and Slow*, Chicago edition

---
class: inverse, middle

# The rise of empirical evidence

---
class: clear-slide

.center[**The fading *American dream***]
```{R, echo = F, dev = "svg", fig.height = 5.5}
data <- read_csv("fading_american_dream.csv")

data %>% 
  ggplot(aes(x = cohort, y = cohort_mean)) +
  geom_point(size = 2, color = red_pink) +
  geom_line(size = 1, color = red_pink) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Child's birth cohort", y = "Children earning more than their parents") +
  theme_simple
```

.footnote[*Source:* Raj Chetty et al. (2017), [The fading American dream: Trends in absolute income mobility since 1940](https://science.sciencemag.org/content/356/6336/398.abstract), *Science*.]

---
# Why is the *American dream* fading?

**Policy Question:** Why is a child's chance of climbing the income ladder decreasing in the United States?

- What can we do to reverse this trend?

Difficult to answer with historical data on macroeconomic trends.

- **That other things change over time makes it difficult** to separately identify the roles of alternative explanations

---
# Theoretical social science

Historically, the social sciences had **limited data** to study policy questions.

**The result?** Social sciences were .pink[theoretical] fields

- Some researchers developed .pink[mathematical models]
- Some developed .pink[qualitative theories]
- Both used their theories to make policy recommendations (*e.g.,* to improve upward mobility)

--

**The problem?** Without data, even falsifiable theories are never tested!

- Five researchers could have five different answers to the same question
- Can lead to a politicization of questions that, in principle, have scientific answers (*e.g.,* do minimum wage laws reduce employment?)

---
class: clear-slide

.center[**Economics is becoming more data-driven** <br>.slate[Empirical articles in the top three economics journals over time]]
```{R, echo = F, dev = "svg", fig.height = 5}
data <- tibble(
  percent = c(38.4, 60.3, 60.0, 72.1),
  year = c("1983", "1993", "2003", "2011")
)

data %>% 
  ggplot(aes(x = year, y = percent, label = percent)) +
  scale_y_continuous(breaks = seq(0, 80, 10), expand = c(0, 0)) +
  geom_col(size = 2, fill = red_pink) +
  geom_text(size = 5, position = position_stack(vjust = 0.5), color = "white", family = "Fira Sans") +
  labs(y = "Empirical articles (% of total)") +
  theme_simple +
  theme(axis.title.x = element_blank())
```

.footnote[*Source:* Daniel S. Hamermesh (2013), [Six Decades of Top Economics Publishing: Who and How?](https://www.aeaweb.org/articles?id=10.1257/jel.51.1.162) *Journal of Economic Literature*.]

---
# The rise of empirical evidence

Today, the social sciences are increasingly **empirical** thanks to the growing availability of data and computational power.

- Gives us **the ability to test** existing theories
- Gives us **the ability to refine** theory to (i) better explain decision making and (ii) better fit real-world data

--

The social sciences have caught up to the natural sciences in terms of scientific rigor, arguably surpassing the natural sciences in sophistication.

- Given the complexity of human decision making, inability to experiment in controlled environments, *etc.*?

---
class: inverse, middle

# Making *other-things-equal* comparisons

---
# *Other-things-equal* comparisons

**The policy?** In 2017, the University of Oregon started requiring first-year students to live on campus.

**The rationale?** First-year students who live on campus outperform those who live off campus.

- Average 2.super[nd]-year retention rate _5 percentage points higher_
- _80 percent more likely to graduate_ in four years
- GPA _0.13 points higher_

--

**Q:** Do these comparisons suggest that the policy will improve student outcomes?

--

**Q:** Do they describe the effect of living on campus?

**Q:** Do they describe *something else?*

---
# *Other-things-equal* comparisons

**Healthy skepticism** should leave us questioning the UO's interpretation.

- The **decision** to live on campus is likely related to family wealth and interest in school.
- Family wealth and interest in school are also related to academic achievement.


--



**The difference in outcomes** between those on and off campus **does not offer an *other-things-equal* comparison**.

- Without further evidence, one should not attribute the difference in outcomes to living on campus.
    - Not without considering those things that both (i) correlate with living on campus (*e.g.,* family wealth) and (ii) correlate with outcomes (*e.g.,* graduation)

---
# *Other-things-equal* comparisons

Statistical comparisons can only identify causal relationships between variables **when all other factors are "held constant."** 

- *Causal* relationship .mono[=] How a change in one variable *induces* a change in another

--

Economists have developed a *comparative advantage*.super[.hi-pink[<span>&#8224;</span>]] in understanding where **other-things-equal** comparisons can (and cannot) be made.

.footnote[.super[.hi-pink[<span>&#8224;</span>]] *Comparative advantage* .mono[=] Ability of an individual or group to perform an activity at lower cost relative to another individual or group.]

- Anyone can retort "_correlation doesn't imply causation!_"
--

- Understanding why it doesn't? The conditions under which it actually does imply causality?
    - Difficult, but necessary for learning from data!

---
class: inverse, middle

# What do "data" look like?

---
# What do "data" look like?

.more-left[

```{r, echo=FALSE}
data(wage1)
wage1 <- select(wage1, wage, educ, female) %>%
  mutate(wage = round(wage, 2))
DT::datatable(
  wage1,
  caption = c("Sample of US workers (1976)"),
  colnames = c('<span style="color: #007935 !important">Wage</span>', '<span style="color: #007935 !important">Education</span>', '<span style="color: #007935 !important">Female?</span>'),
  fillContainer = FALSE, options = list(pageLength = 5, lengthChange = FALSE, searching = FALSE), escape = FALSE) %>%
  # DT::formatStyle('wage', color = '#9370DB') %>%
  # DT::formatStyle('educ', color = '#9370DB') %>%
  # DT::formatStyle('female', color = '#9370DB') %>%
  DT::formatStyle(0, color = '#e64173')
```

]

.less-right[

.hi-pink[Rows] represent .hi-pink[observations].

.hi-green[Columns] represent .hi-green[variables].

Each value is associated with an .hi-pink[observation] and a .hi-green[variable].

]


---
class: inverse, middle

# Causal identification

---
# Causal identification

## The objective

Identify the effect of a .hi[treatment] on an .hi[outcome].

--

## The ideal comparison

Ideally, we could calculate the **treatment effect** *for each individual* as

$$Y_{1,i} - Y_{0,i}$$

- $Y_{1,i}$ is the outcome for person $i$ when $i$ receives the treatment
- $Y_{0,i}$ is the outcome for person $i$ when $i$ does not receive the treatment
- Known as **potential outcomes**

---
# Causal identification

The **ideal data** for 10 people

.pull-left[
```{R, ideal_data, echo = F}
set.seed(3)
ideal_df <- data.frame(
  i = 1:10,
  treat = rep(c(1, 0), each = 5),
  Y_1i = c(runif(10, 4, 10) %>% round(2)),
  Y_0i = c(runif(10, 0, 5) %>% round(2))
) %>% 
  mutate(Y_0i = case_when(treat == 1 ~ Y_0i + 2, TRUE ~ Y_0i),
         Y_1i = case_when(treat == 0 ~ Y_1i, TRUE ~ Y_1i))
ideal_df
```
]

--

.pull-right[
We could calculate the treatment effect for each individual $i$,
$$
\begin{align}
  \tau_i = Y_{1,i} -  Y_{0,i}~,
\end{align}
$$
and we would be inclined to think of it as the causal effect.
]

---
count: false
# Causal identification

The **ideal data** for 10 people

.pull-left[
```{R, ideal_data_trt, echo = F}
ideal_df %>% mutate(effect_i = Y_1i - Y_0i)
```
]

.pull-right[
We could calculate the treatment effect for each individual $i$,
$$
\begin{align}
  \tau_i = Y_{1,i} -  Y_{0,i}~,
\end{align}
$$
and we would be inclined to think of it as the causal effect.
]

---
count: false
# Causal identification

The **ideal data** for 10 people

.pull-left[
```{R, ideal_data_trt2, echo = F}
ideal_df %>% mutate(effect_i = Y_1i - Y_0i)
```
]

.pull-right[
We could calculate the treatment effect for each individual $i$,
$$
\begin{align}
  \tau_i = Y_{1,i} -  Y_{0,i}~,
\end{align}
$$
and we would be inclined to think of it as the causal effect.

]

The mean of these individual treatment effects .mono[=] `r transmute(ideal_df, effect_i = Y_1i - Y_0i) %>% unlist %>% mean %>% round(2)`

- We call this the .hi-green[average treatment effect] (ATE)

---
# Causal identification

## The fundamental problem of causal inference

While the ideal comparison is
$$
\begin{align}
  \tau_i = \color{#e64173}{Y_{1,i}} &- \color{#9370DB}{Y_{0,i}}~,
\end{align}
$$
this comparison is fundamentally challenged!

--

- If we observe $\color{#e64173}{Y_{1}}$ for $i$, then we cannot observe $\color{#9370DB}{Y_{0}}$ for $i$
- If we observe $\color{#9370DB}{Y_{0}}$ for $i$, then we cannot observe $\color{#e64173}{Y_{1}}$ for $i$
- We only observe what *actually* happened&mdash;we cannot observe the **counterfactual**

--

**The implication?** .hi-pink[ALL] .pink[causal inference is] .hi-pink[by assumption!]

---
# Causal identification

The data we *actually* see for these 10 people?
.pull-left[
```{R, ideal_data_obs, echo = F}
obs_df <- ideal_df
obs_df$Y_0i[1:5] <- NA
obs_df$Y_1i[6:10] <- NA
obs_df
```
]

--

.pull-right[
We only observe $\color{#e64173}{Y_{1}}$ for $i \in \{1, ..., 5\}$

We only observe $\color{#9370DB}{Y_{0}}$ for $i \in \{6, ..., 10\}$
 
We do not observe both $\color{#e64173}{Y_{1,i}}$ and $\color{#9370DB}{Y_{0,i}}$ for anyone

]

--

**Q:** How can we estimate the average treatment effect when we cannot observe individual treatment effects?

---
# Causal identification

Can we **compare the mean outcomes** of each group?

- Take the average of $\color{#e64173}{Y_{1}}$ for those who received the treatment (*i.e.,* the .pink[treatment-group mean])
- Take the average of $\color{#9370DB}{Y_{0}}$ for those who didn't receive the treatment (*i.e.,* the .purple[control-group mean])

--

**Q:** Does .pink[treatment-group mean] .mono[-] .purple[control-group mean] isolate the causal effect of the treatment?

---
# Causal identification

.pull-left[
```{R, ideal_data_obs_2, echo = F}
obs_df <- ideal_df
obs_df$Y_0i[1:5] <- NA
obs_df$Y_1i[6:10] <- NA
obs_df
```
]

.pull-right[
.pink[Treatment group mean] .mono[=] `r round(mean(obs_df$Y_1i[1:5]), 2)`

.purple[Control group mean] .mono[=] `r round(mean(obs_df$Y_0i[6:10]), 2)`

Difference-in-means .mono[=] `r round(mean(obs_df$Y_1i[1:5]) - mean(obs_df$Y_0i[6:10]), 2)`
]

--

Difference-in-means .mono[=] .hi-green[average treatment effect] .mono[+] .hi-orange[selection bias]
--
<br> $\quad$ .mono[=] .green[`r transmute(ideal_df, effect_i = Y_1i - Y_0i) %>% unlist %>% mean %>% round(2)`] .mono[+] .orange[(`r round(mean(obs_df$Y_1i[1:5]) - mean(obs_df$Y_0i[6:10]), 2)` .mono[-] `r transmute(ideal_df, effect_i = Y_1i - Y_0i) %>% unlist %>% mean %>% round(2)`)]
--
 .mono[=] .green[`r transmute(ideal_df, effect_i = Y_1i - Y_0i) %>% unlist %>% mean %>% round(2)`] .mono[+] .orange[`r round(mean(obs_df$Y_1i[1:5]) - mean(obs_df$Y_0i[6:10]), 2) - transmute(ideal_df, effect_i = Y_1i - Y_0i) %>% unlist %>% mean %>% round(2)`]

--

.orange[Selection bias] .mono[!=] 0 .mono[==>] people who "select into" treatment are different

---
class: inverse, middle

# Randomized control trials 

---
# Randomized control trials 

## Overcoming selection bias

**The problem?** The existence of selection bias precludes making *other-things-equal* comparisons.

- To make valid comparisons that identify causal effects, we need to shut down the bias coming from selection.

--

**The solution?** Conduct an experiment!

- How? Assign treatment .hi-pink[randomly]
- Hence the name, .hi-pink[*randomized* control trial] (RCT)

---
# Randomized control trials 

## Example: Effect of de-worming on attendance

**Motivation:** Intestinal worms are common among children in less-developed countries. The symptoms of these parasites can keep school-aged children at home, disrupting human capital accumulation.

**Policy question:** Do school-based de-worming interventions provide a cost-effective way to increase school attendance? 

---
# Randomized control trials 

## Example: Effect of de-worming on attendance

**Research question:** How much do de-worming interventions increase school attendance?

**Q:** **Could we simply compare average attendance** among children with and without access to de-worming medication?

--

- **A:** If we're after the causal effect, probably not. (Why not?)


--

**Selection bias:** Families with access to de-worming medication probably have healthier children for other reasons, too (wealth, access to clean drinking water, *etc.*).

- **We can't make an *all-else-equal* comparison** .mono[-->] in expectation, observed differences will deviate *systematically* from the ATE!

---
# Randomized control trials 

## Example: Effect of de-worming on attendance

**Solution:** Run an experiment.

--

Imagine an RCT where we have two groups:

- .hi-slate[Treatment:] Villages where children get de-worming medication in school.
- .hi-slate[Control:] Villages where children don't get de-worming medication in school (status quo).

--

By randomizing villages into .hi-slate[treatment] or .hi-slate[control], we will, on average, include all kinds of villages (poor _vs._ less poor, access to clean water _vs._ contaminated water, hospital _vs._ no hospital, *etc.*) in both groups.

--

*All else equal*!

---
class: clear-slide

.hi-slate[72 villages] 
```{R, plot1, echo = F, dev = "svg", fig.height = 5.5}
xd <- 12
yd <- 6
set.seed(123)
med_df <- expand.grid(x = 1:xd, y = 1:yd) %>%
  mutate(
    q = x + y + 1.25 * rnorm(xd*yd),
    trt = sample(x = c(T, F), size = xd*yd, replace = T, prob = c(0.51, 0.49)),
  ) %>%
  arrange(-y, x) %>%
  mutate(id = 1:(xd*yd) %>% str_pad(2, "left", "0")) %>%
  arrange(y, x)
ggplot(data = med_df, aes(x, y)) +
geom_tile(color = "grey40", fill = "grey93", size = 0.2) +
geom_text(aes(label = id), color = "grey45", size = 5, family = "Roboto Mono") +
theme_void() +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development]
```{R, plot2, echo = F, dev = "svg", fig.height = 5.5}
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_1, echo = F, dev = "svg", fig.height = 5.5}
set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
geom_emoji(data = sample_n(med_df, 1), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_2, echo = F, dev = "svg", fig.height = 5.5}
set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
geom_emoji(data = sample_n(med_df, 2), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_3, echo = F, dev = "svg", fig.height = 5.5}
set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
geom_emoji(data = sample_n(med_df, 3), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_4, echo = F, dev = "svg", fig.height = 5.5}
set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
geom_emoji(data = sample_n(med_df, 4), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_5, echo = F, dev = "svg", fig.height = 5.5}
set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
geom_emoji(data = sample_n(med_df, 5), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_6, echo = F, dev = "svg", fig.height = 5.5}
set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
geom_emoji(data = sample_n(med_df, 6), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_7, echo = F, dev = "svg", fig.height = 5.5}
set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, 7), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_8, echo = F, dev = "svg", fig.height = 5.5}
set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, 8), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_9, echo = F, dev = "svg", fig.height = 5.5}
set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, 9), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_10, echo = F, dev = "svg", fig.height = 5.5}
set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, round(xd * yd / 2, 0)), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_11, echo = F, dev = "svg", fig.height = 5.5}
set.seed(248)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, round(xd * yd / 2, 0)), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---
class: clear-slide
count: false

.hi-slate[72 villages] .hi[of varying levels of development] .mono[+] .hi-orange[randomly assigned treatment]
```{R, plot3_12, echo = F, dev = "svg", fig.height = 5.5}
set.seed(250)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, round(xd * yd / 2, 0)), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---
# Randomized control trials 

## Example: Effect of de-worming on attendance

We can estimate the **causal effect** of de-worming on school attendance by **comparing the average attendance rates** in the .hi-pink[treatment group] (💊) with those in the .hi-purple[control group] (no 💊):

.center[.hi-pink[Treatment group attendance rate] .mono[-] .hi-purple[Control group attendance rate]]


--

**Result:** This was done in Kenya, where [attendance increased](https://www.povertyactionlab.org/case-study/deworming-schools-improves-attendance-and-benefits-communities-over-long-term) with the random assignment of treatment.

- 25-percent decrease in absenteeism at a cost of $0.60 per child
- Long term cost effectiveness: Additional 11.91 years of schooling per $100 spent on de-worming

.footnote[*Source:* [Deworming to increase school attendance](https://www.povertyactionlab.org/case-study/deworming-increase-school-attendance), *Abdul Latif Jameel Poverty Action Lab*.]

---
# Randomized control trials 

## Example: Effect of de-worming on attendance

We can estimate the **causal effect** of de-worming on school attendance by **comparing the average attendance rates** in the .hi-pink[treatment group] (💊) with those in the .hi-purple[control group] (no 💊):

.center[.hi-pink[Treatment group attendance rate] .mono[-] .hi-purple[Control group attendance rate]]

**Q:** Should we trust the results of the comparison? 

--

**A:** Even with healthy skepticism, we probably should? On average, randomly assigning treatment balances the treatment and control groups across other dimensions that could explain school attendance.

---
class: clear-slide

Balance ***on average*** .mono[!=] Balance ***every time***
```{R, fertilizer_plot3_bad, echo = F, dev = "svg", fig.height = 5.5}
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = filter(med_df, x < xd/3 | y <= yd/3), emoji = "1f48a") + #  | (x == xd & y == yd)
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---
# Interpreting results

## Internal validity

Addresses the question, ***should we believe the study?***

A study has high **internal validity** if, within the context of the study, we are confident that one variable has a **causal** influence on the outcome of interest (*e.g.,*  there's **no selection bias**).

--

## External validity

Addresses the question, ***how far can we generalize the results of the study?***

A study has high **external validity** to the extent that the results **apply to other contexts** (not just the local environment that generated the results).

---
class: inverse, middle

# *Thinking Fast and Slow*, Chicago edition

---
# *Thinking Fast and Slow*, Chicago edition

## Background

**Policy question:** How can we reduce violent crime among young men? 

**Research agenda:** What factors influence an individual's proclivity toward violent crime? 

- Self control? Social skills? Grit?
- Economic hardship?
- Police presesnce?
- Early chilhood education?
- Something else?

---
# *Thinking Fast and Slow*, Chicago edition

**Research question:** Can cognitive-behavioral therapy keep young men in school and out of trouble?

- Proposed mechanism: Automaticity.

.footnote[*Source:* Sara B Heller et al. (2017), [Thinking, Fast and Slow? Some Field Experiments to Reduce Crime and Dropout in Chicago](https://academic.oup.com/qje/article-abstract/132/1/1/2724542?redirectedFrom=fulltext), *The Quarterly Journal of Economics*.]

--

**Experiment:** *Becoming a Man*

4804 young men in Chicago Public Schools randomly assigned to one of two groups:

- .hi-pink[Treatment group:] Group cognitive-behavioral therapy program during school (once per week for 1-2 school years)
- .hi-purple[Control group:] No intervention

--

A similar experiment was also conducted in the Cook County Juvenile Temporary Detention Center.

---
class: clear-slide

.center[***Becoming a Man*: Experimental results**]
```{R, echo = F, escape = F}
tab <- data.frame(
  v1 = c(
    "School engagement index",
    "",
    "Total arrests per youth per year",
    "",
    "&nbsp;&nbsp;&nbsp; Violent",
    "",
    "&nbsp;&nbsp;&nbsp; Property",
    "",
    "&nbsp;&nbsp;&nbsp; Drug",
    "",
    "&nbsp;&nbsp;&nbsp; Other",
    ""
  ),
  v2 = rbind(
    c(0, 0.603, 0.136, 0.069, 0.132, 0.266),
    c("", "", "", "", "", "")
  ) %>% as.vector(),
  v3 = rbind(
    c(0.040, 0.53, 0.109, 0.072, 0.127, 0.222),
    c("", "", "", "", "", "")
  ) %>% as.vector(),
  v4 = rbind(
    c(0.040, -0.073, -0.027, 0.003, -0.005, -0.044),
    c("(0.016)", "(0.031)", "(0.011)", "(0.008)", "(0.012)", "(0.019)")
  ) %>% as.vector(),
  v5 = rbind(
    c(0.088, -0.161, -0.060, 0.006, -0.011, -0.097),
    c("(0.034)", "(0.068)", "(0.024)", "(0.018)", "(0.027)", "(0.040)")
  ) %>% as.vector()
) %>% kable(
  escape = F,
  col.names = c(
    "Outcome",
    "Control mean",
    "Treatment mean",
    "Effect of treatment assignment",
    "Effect of participation"
  ),
  align = c("l", rep("c", 4)),
) %>%
  row_spec(1:12, color = met_slate) %>%
  row_spec(seq(2,12,2), color = "#c2bebe") %>%
  row_spec(1:12, extra_css = "line-height: 110%;") %>%
  column_spec(1, color = "black", italic = T) %>% 
  kable_styling(font_size = 20.5, bootstrap_options = "hover") #%>% 
  # footnote(general = "4804 observations from study 1 and study 2 pooled together. Standard errors in parentheses.",
  #          general_title = "Notes:",
  #          title_format = "italic",
  #          footnote_as_chunk = T)
tab
```
.smallest[*Notes:* 4804 observations. Standard errors in parentheses.]

---
# Housekeeping

**Assigned reading for ~~Wednesday~~ Friday:** [Snapping back: Food stamp bans and criminal recidivism](https://www.aeaweb.org/articles?id=10.1257/pol.20170490) by Cody Tuttle (2019).

- I will post the paper on Canvas tonight!
- Reading Quiz 3 will be due this Friday.
- Details to follow!