---
title: "Learning from Observational Data"
subtitle: "EC 350: Labor Economics"
author: "[Kyle Raze](https://kyleraze.com)"
date: "Winter 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
      ratio: "16:9"
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(ggthemes, readxl, viridis, broom, emoGG, ggdag, gganimate, knitr, dslabs, gapminder, extrafont, Ecdat, wooldridge, tidyverse, magrittr, janitor, kableExtra, gridExtra, ggforce)
# Define colors
red_pink <- "#e64173"
met_slate <- "#272822" # metropolis font color 
purple <- "#9370DB"
green <- "#007935"
light_green <- "#7DBA97"
orange <- "#FD5F00"
turquoise <- "#44C1C4"
red <- "#b92e34"
slate <- "#314f4f"
# Notes directory
dir_slides <- "~/GitHub/Intro_Labor_S21/Lectures/04-Observational_Data/"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  #dpi = 300,
  #cache = T,
  warning = F,
  message = F
)  
theme_simple <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  # axis.text.x = element_text(size = 12),
  # axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
theme_market <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.title.x = element_text(hjust = 1, size = 17),
  axis.title.y = element_text(hjust = 1, angle = 0, size = 17),
  # axis.text.x = element_text(size = 12),
  # axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
theme_gif <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
wrapper <- function(x, ...) paste(strwrap(x, ...), collapse = "\n")
# police data
data <- get(data(campus)) %>% 
  mutate(crime = round(crime/enroll*1000, 2),
         police = round(police/enroll*1000, 2)) %>% 
  filter(police < 10) %>% # remove outlier
  select(crime, police)
lm0 <- lm(crime ~ police, data = data)
wage2 <- get('wage2')
lm_wage <- lm(wage ~ educ, wage2)

experimental <- get("jtrain2")
observational <- get("jtrain3")

fixest::feols(unem78 ~ train + csw0(age + black + hisp, educ, unem74), data = experimental) %>%
  fixest::etable(digits = 3)
p_load(pander)
fixest::feols(unem78 ~ train + csw0(age + agesq + black + hisp, educ, unem74), data = observational) %>%
  fixest::etable(digits = 3)
```

# Prologue

---
# Learning from Observational Data

1. A taxonomy of data
    - Experimental *vs.* observational data
2. Direct acyclic graphs
    - Causal paths
    - Backdoor paths
    - Backdoor criterion
3. Regression discontinuity

---
class: inverse, middle

# A taxonomy of data

---
# A taxonomy of data

## .hi-pink[Experimental]

Data generated from a .hi-pink[randomized] experiment.

- Treatment assigned at .hi-pink[random]
- The **gold standard** of social science research
- Often difficult/impractical/unethical to conduct

--

## .hi-purple[Observational (non-experimental)]

Data generated from the .hi-purple[decisions] of various individuals in the "real world."

- Sometimes treatment is randomly assigned (*e.g.,* in a lottery), but not usually .hi-purple[(non-random!)]
- Prone to selection bias
- Must rely on natural experiments to identify causal relationships

---
# A taxonomy of data

## **Example: Effect of job training on unemployment status**

--

.pull-left[
### .hi-pink[Experimental sample] 

```{css, echo=F}
    /* Table width = 100% max-width */

    .remark-slide table{
        width: 100%;
    }

    /* Change the background color to white for shaded rows (even rows) */

    .remark-slide thead, .remark-slide tr:nth-child(2n) {
        background-color: white;
    }
```

```{R, echo = F, escape = F}
tab <- data.frame(
  v1 = c(".hi-pink[Training?]" , "", "**Demographics**", "**Education**", "**Previously unemployed?**"),
  v2 = c("-0.111", "(0.044)", "", "", ""),
  v3 = c("-0.116", "(0.044)", "$\\checkmark$", "", ""),
  v4 = c("-0.115", "(0.044)", "$\\checkmark$", "$\\checkmark$", ""),
  v5 = c("-0.113", "(0.044)", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$")
) %>% kable(
  escape = F,
  col.names = c("", "1", "2", "3", "4"),
  align = c("l", rep("c", 5)),
  caption = "<br>**Unemployed?** (.mono[=] 1 if yes, .mono[=] if no)"
) %>%
row_spec(0, color = "#708090") %>% 
row_spec(1:5, color = met_slate) %>%
row_spec(2, color = "#c2bebe") %>%
column_spec(1, color = met_slate)

tab
```
.center[*Note:* Standard errors in parentheses.]
]

--

.pull-right[
### .hi-purple[Non-experimental sample] 

```{css, echo=F}
    /* Table width = 100% max-width */

    .remark-slide table{
        width: 100%;
    }

    /* Change the background color to white for shaded rows (even rows) */

    .remark-slide thead, .remark-slide tr:nth-child(2n) {
        background-color: white;
    }
```

```{R, echo = F, escape = F}
tab <- data.frame(
  v1 = c(".hi-purple[Training?]" , "", "**Demographics**", "**Education**", "**Previously unemployed?**"),
  v2 = c("0.128", "(0.025)", "", "", ""),
  v3 = c("0.164", "(0.027)", "$\\checkmark$", "", ""),
  v4 = c("0.160", "(0.027)", "$\\checkmark$", "$\\checkmark$", ""),
  v5 = c(-0.182, "(0.027)", "$\\checkmark$", "$\\checkmark$", "$\\checkmark$")
) %>% kable(
  escape = F,
  col.names = c("", "1", "2", "3", "4"),
  align = c("l", rep("c", 5)),
  caption = "<br>**Unemployed?** (.mono[=] 1 if yes, .mono[=] if no)"

) %>%
row_spec(0, color = "#708090") %>% 
row_spec(1:5, color = met_slate) %>%
row_spec(2, color = "#c2bebe") %>%
column_spec(1, color = met_slate)

tab
```
.center[*Note:* Standard errors in parentheses.]

]


---
class: inverse, middle

# Direct acyclic graphs

---
# Direct acyclic graphs

.pull-left[

A direct acyclic graph (DAG) can help us visualize the assumptions necessary to estimate causal relationships using observational data.

.hi-pink[Nodes] represent .hi-pink[variables].

.hi-black[Arrows] represent .hi-black[causal relationships] between variables.

]
.pull-right[
```{r, dev = "svg", echo = F}
dagify(Y ~ D + X,
       D ~ X,
       exposure = "D",
       outcome = "Y") %>%
  ggplot(aes(
    x = x,
    y = y,
    xend = xend,
    yend = yend
  )) +
  geom_dag_point(color = red_pink) +
  geom_dag_edges_link(arrow = grid::arrow(length = grid::unit(10, "pt"), type = "closed")) +
  geom_dag_text(family = "Fira Sans", size = 10) +
  theme_dag()
```

]

---
# DAGs follow two rules
.pull-left[
.center[**Rule 1 ("direct"):** No bidirectional arrows!]


```{r, dev = "svg", echo = F}
dagify(Y ~ D,
       D ~ Y) %>%
  ggplot(aes(
    x = x,
    y = y,
    xend = xend,
    yend = yend
  )) +
  geom_dag_point(color = red_pink) +
  geom_dag_edges_link(arrow = grid::arrow(length = grid::unit(10, "pt"), type = "closed")) +
  geom_dag_text(family = "Fira Sans", size = 10) +
  theme_dag()
```
.center[.hi-red[Illegal!]]
]
--
.pull-right[
.center[**Rule 2 ("acyclic"):** No feedback loops!]

```{r, dev = "svg", echo = F}
dagify(Y ~ D,
       X ~ Y,
       D ~ X,
       exposure = "D",
       outcome = "Y") %>%
  ggplot(aes(
    x = x,
    y = y,
    xend = xend,
    yend = yend
  )) +
  geom_dag_point(color = red_pink) +
  geom_dag_edges_link(arrow = grid::arrow(length = grid::unit(10, "pt"), type = "closed")) +
  geom_dag_text(family = "Fira Sans", size = 10) +
  theme_dag()
```
.center[.hi-red[Illegal!]]
]

---
# Causal paths

Our objective is to **identify the causal effect** of a treatment variable .hi[D] on an outcome variable .hi[Y].

- The treatment could have a **direct effect** on the outcome: .hi[D] .mono[-->] .hi[Y].
- Alternatively, the treatment could have an **indirect effect** on the outcome through .hi[X], a mediator variable: .hi[D] .mono[-->] .hi[X] .mono[-->] .hi[Y].

```{r, dev = "svg", echo = F, fig.height = 3.5}
dagify(Y ~ D + X,
       X ~ D,
       exposure = "D",
       outcome = "Y") %>%
  ggplot(aes(
    x = x,
    y = y,
    xend = xend,
    yend = yend
  )) +
  geom_dag_point(color = red_pink) +
  geom_dag_edges_link(arrow = grid::arrow(length = grid::unit(5, "pt"), type = "closed")) +
  geom_dag_text(family = "Fira Sans", size = 5) +
  theme_dag()
```

---
# Backdoor paths

The presence of a confounder variable .hi[W] opens a **backdoor path** from the treatment to the outcome:<br>.center[.hi[D] .mono[<--] .hi[W] .mono[-->] .hi[Y]]

```{r, dev = "svg", echo = F, fig.height = 3.5}
dagify(Y ~ D + W,
       D ~ W,
       exposure = "D",
       outcome = "Y") %>%
  ggplot(aes(
    x = x,
    y = y,
    xend = xend,
    yend = yend
  )) +
  geom_dag_point(color = red_pink) +
  geom_dag_edges_link(arrow = grid::arrow(length = grid::unit(5, "pt"), type = "closed")) +
  geom_dag_text(family = "Fira Sans", size = 5) +
  theme_dag()
```

An open backdoor path creates a **spurious correlation** between the treatment and the outcome!

---
# Backdoor paths

## **Example: Returns to education**

.pull-left[
**Q:** How does education affect earnings?

- .hi[D] .mono[=] Education (*e.g.,* going to college or not)
- .hi[Y] .mono[=] Earnings as an adult
- .hi[PE] .mono[=] Parental education
- .hi[I] .mono[=] Family income
- .hi[U] .mono[=] Unobserved characteristics (*e.g.,* family background)

]
.pull-right[
```{r, dev = "svg", echo = F}
returns_to_ed <-
  dagify(Y ~ D + I,
         I ~ PE,
         PE ~ U,
         D ~ U,
         D ~ PE,
         D ~ I,
         exposure = "D",
         outcome = "Y") %>%
  ggplot(aes(
    x = x,
    y = y,
    xend = xend,
    yend = yend
  )) +
  geom_dag_point(color = red_pink) +
  geom_dag_edges_link(arrow = grid::arrow(length = grid::unit(10, "pt"), type = "closed")) +
  geom_dag_text(family = "Fira Sans", size = 10) +
  theme_dag()
returns_to_ed
```
]

---
count: false
# Backdoor paths

## **Example: Returns to education**

.pull-left[
**Q:** How does education affect earnings?

- .hi[D] .mono[=] Education (*e.g.,* going to college or not)
- .hi[Y] .mono[=] Earnings as an adult
- .hi[PE] .mono[=] Parental education
- .hi[I] .mono[=] Family income
- .hi[U] .mono[=] Unobserved characteristics (*e.g.,* family background)

The presence&mdash;*or absence*&mdash;of an arrow illustrates our **causal assumptions** about how education affects earnings!
]
.pull-right[
```{r, dev = "svg", echo = F}
returns_to_ed
```
]

---
# Backdoor paths

## **Example: Returns to education**

.pull-left[
**Q:** What are the paths through which education affects earnings?

]
.pull-right[
```{r, dev = "svg", echo = F}
returns_to_ed
```
]

---
count: false
# Backdoor paths

## **Example: Returns to education**

.pull-left[
**Q:** What are the paths through which education affects earnings?

- .hi[D] .mono[-->] .hi[Y] (causal effect)
- .hi[D] .mono[<--] .hi[I] .mono[-->] .hi[Y] (backdoor path)
- .hi[D] .mono[<--] .hi[PE] .mono[-->] .hi[I] .mono[-->] .hi[Y] (backdoor path)
- .hi[D] .mono[<--] .hi[U] .mono[-->] .hi[PE] .mono[-->] .hi[I] .mono[-->] .hi[Y] (backdoor path)

]
.pull-right[
```{r, dev = "svg", echo = F}
returns_to_ed
```
]

---
# Backdoor paths

## **Backdoor criterion**

> The observed correlation between .hi[Y] and .hi[D] isolates the causal effect of .hi[D] on .hi[Y] if and only if all backdoor paths from .hi[D] to .hi[Y] are closed.

--

**Q:** What closes a backdoor path?

- **A.sub[1]:** *Conditioning* or *controlling for* the confounder variable on the path.
--

- **A.sub[2]:** The presence of a collider variable on the path.


---
# Backdoor paths 

The presence of a collider variable .hi[C] closes a backdoor path from the treatment to the outcome:<br>.center[.hi[D] .mono[-->] .hi[C] .mono[<--] .hi[Y]]

```{r, dev = "svg", echo = F, fig.height = 3.25}
dagify(Y ~ D,
       C ~ Y,
       C ~ D,
       exposure = "D",
       outcome = "Y") %>%
  ggplot(aes(
    x = x,
    y = y,
    xend = xend,
    yend = yend
  )) +
  geom_dag_point(color = red_pink) +
  geom_dag_edges_link(arrow = grid::arrow(length = grid::unit(5, "pt"), type = "closed")) +
  geom_dag_text(family = "Fira Sans", size = 5) +
  theme_dag()
```

--

**The implication?** We don't want to control for collider variables!

- Conditioning on a collider can open up new backdoor paths. (More on this later.)

---
# Backdoor paths

## **Example: Returns to education**

.pull-left[
**Q:** How could we satisfy the backdoor criterion given our assumptions about the effect of education on earnings?

]
.pull-right[
```{r, dev = "svg", echo = F}
returns_to_ed
```
]

---
count: false
# Backdoor paths

## **Example: Returns to education**

.pull-left[
**Q:** How could we satisfy the backdoor criterion given our assumptions about the effect of education on earnings?

**A:** Control for family income (.hi[I])

- **Why?** Family income appears as a non-collider on each backdoor path:<br>.center[.hi[D] .mono[<--] .hi[I] .mono[-->] .hi[Y]] .center[.hi[D] .mono[<--] .hi[PE] .mono[-->] .hi[I] .mono[-->] .hi[Y]] .center[.hi[D] .mono[<--] .hi[U] .mono[-->] .hi[PE] .mono[-->] .hi[I] .mono[-->] .hi[Y]]

]
.pull-right[
```{r, dev = "svg", echo = F}
returns_to_ed
```
]

---
# Backdoor paths

## **Example: Returns to education**

.pull-left[
**Q:** Would controlling for family income isolate the causal effect of education on earnings if unobserved family background (.hi[U]) has a direct effect on earnings (.hi[Y])?

]
.pull-right[
```{r, dev = "svg", echo = F}
returns_to_ed_2 <-
  dagify(Y ~ D + I,
         I ~ PE,
         PE ~ U,
         D ~ U,
         D ~ PE,
         D ~ I,
         Y ~ U,
         exposure = "D",
         outcome = "Y") %>%
  ggplot(aes(
    x = x,
    y = y,
    xend = xend,
    yend = yend
  )) +
  geom_dag_point(color = red_pink) +
  geom_dag_edges_link(arrow = grid::arrow(length = grid::unit(10, "pt"), type = "closed")) +
  geom_dag_text(family = "Fira Sans", size = 10) +
  theme_dag()
returns_to_ed_2
```
]

---
count: false
# Backdoor paths

## **Example: Returns to education**

.pull-left[
**Q:** Would controlling for family income isolate the causal effect of education on earnings if unobserved family background (.hi[U]) has a direct effect on earnings (.hi[Y])?

**A:** No! 

- .hi[U] is unobserved, so we can't control for it.
- The backdoor path .hi[D] .mono[<--] .hi[U] .mono[-->] .hi[Y] would stay open.
]
.pull-right[
```{r, dev = "svg", echo = F}
returns_to_ed_2
```
]

---
count: false
# Backdoor paths

## **Example: Returns to education**

.pull-left[
**Q:** Would controlling for family income isolate the causal effect of education on earnings if unobserved family background (.hi[U]) has a direct effect on earnings (.hi[Y])?

**A:** No! 

- .hi[U] is unobserved, so we can't control for it.
- The backdoor path .hi[D] .mono[<--] .hi[U] .mono[-->] .hi[Y] would stay open.

**The takeaway?**<br>.hi-pink[ALL causal inference is by assumption!]
]
.pull-right[
```{r, dev = "svg", echo = F}
returns_to_ed_2
```
]

---
class: inverse, middle

# Regression discontinuity

---
# Regression discontinuity

There are situations in the real world where treatment is assigned in a way that is **as good as random.** 

- These situations can provide **valid comparison groups**, just like the ones you'd find in a randomized control trial!

**Examples?** When some arbitrary threshold triggers a change in treatment: 

- Anti-discrimination laws only apply to firms with more than 15 employees.
- Prisoners are eligible for early parole if some score exceeds a threshold.
- An individual has legal access to alcohol if they are 21 or older.
- You get a ticket if your speed exceeds the speed limit.
- A candidate for governor wins if her vote share exceeds that of her competitors.


--

Economists can (and often do) use these situations to estimate causal effects.

---
# Regression discontinuity

**Example:** Effect of merit scholarships on graduation

- Outcome variable .mono[=] probability of graduation
- Treatment .mono[=] scholarship money
- "Assignment variable" .mono[=] admissions test score (*e.g.,* the SAT)
- "Cutoff/threshold"  .mono[=] minimum score for getting a scholarship (*e.g.,* SAT score of 1200 or higher)

--

**Assumption:** Students *just below* the cutoff are comparable to those *just above* the cutoff.

---
layout: true
class: clear-slide

---
Let's start with potential graduation rates: $\color{#9370DB}{\mathop{E}\left[ \text{Y}_{0,i} \mid \text{SAT}_{i} \right]}$

```{r, s0, include = F}
# Define functions
y0 <- function(x) 0.7 / (1 + exp(-3 * x))
y1 <- function(x) 0.85 / (1 + exp(-4 * x)) + 0.1
gg_df <- data.frame(x = c(-1, 1))
```

```{r, s1, echo = F, cache = T, dev = "svg", fig.height = 5.5}
ggplot(data = gg_df, aes(x)) +
stat_function(fun = y0, color = purple, size = 1) +
scale_x_continuous(
  "SAT score",
  lim = c(-1,1),
  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = c(800, 1000, 1200, 1400, 1600)
) +
scale_y_continuous(
  "Probability of graduating",
  lim = c(0,1),
  labels = scales::percent_format(accuracy = 1)
) +
theme_simple
```
---
count: false
Let's start with potential graduation rates: $\color{#9370DB}{\mathop{E}\left[ \text{Y}_{0,i} \mid \text{SAT}_{i} \right]}$ and $\color{#e64173}{\mathop{E}\left[ \text{Y}_{1,i} \mid \text{SAT}_{i} \right]}$.

```{r, s2, echo = F, cache = T, dev = "svg", fig.height = 5.5}
ggplot(data = gg_df, aes(x)) +
stat_function(fun = y0, color = purple, size = 1) +
stat_function(fun = y1, color = red_pink, size = 1) +
scale_x_continuous(
  "SAT score",
  lim = c(-1,1),
  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = c(800, 1000, 1200, 1400, 1600)
) +
scale_y_continuous(
  "Probability of graduating",
  lim = c(0,1),
  labels = scales::percent_format(accuracy = 1)
) +
theme_simple
```
---
You only get a scholarship if if your .hi-slate[SAT score exceeds the cutoff score].

```{r, s3, echo = F, cache = T, dev = "svg", fig.height = 5.5}
ggplot() +
geom_vline(xintercept = 0, color = slate, size = 1, alpha = 0.75) +
stat_function(
  data = data.frame(x = c(-1,0)), aes(x),
  fun = y0, color = purple, size = 1,
  linetype = "solid",
  xlim = c(-1,0)
) +
stat_function(
  data = data.frame(x = c(0,1)), aes(x),
  fun = y0, color = purple, size = 1,
  linetype = "dotted",
  xlim = c(0,1)
) +
stat_function(
  data = data.frame(x = c(-1,0)), aes(x),
  fun = y1, color = red_pink, size = 1,
  linetype = "dotted",
  xlim = c(-1,0)
) +
stat_function(
  data = data.frame(x = c(0,1)), aes(x),
  fun = y1, color = red_pink, size = 1,
  linetype = "solid",
  xlim = c(0,1)
) +
scale_x_continuous(
  "SAT score",
  lim = c(-1,1),
  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = c(800, 1000, 1200, 1400, 1600)
) +
scale_y_continuous(
  "Probability of graduating",
  lim = c(0,1),
  labels = scales::percent_format(accuracy = 1)
) +
theme_simple
```
---
$\color{#e64173}{\mathop{E}\left[ \text{Y}_{1,i} \mid \text{SAT}_{i} = 1200 \right]} - \color{#9370DB}{\mathop{E}\left[ \text{Y}_{0,i} \mid \text{SAT}_{i} = 1200 \right]}$ gives the .hi-orange[causal effect] .hi-slate[at the cutoff].

```{r, s4, echo = F, cache = T, dev = "svg", fig.height = 5.5}
ggplot() +
geom_vline(xintercept = 0, color = slate, size = 1, alpha = 0.3) +
stat_function(
  data = data.frame(x = c(-1,0)), aes(x),
  fun = y0, color = purple, size = 1,
  linetype = "solid",
  xlim = c(-1,0)
) +
stat_function(
  data = data.frame(x = c(0,1)), aes(x),
  fun = y0, color = purple, size = 1,
  linetype = "dotted",
  xlim = c(0,1)
) +
stat_function(
  data = data.frame(x = c(-1,0)), aes(x),
  fun = y1, color = red_pink, size = 1,
  linetype = "dotted",
  xlim = c(-1,0)
) +
stat_function(
  data = data.frame(x = c(0,1)), aes(x),
  fun = y1, color = red_pink, size = 1,
  linetype = "solid",
  xlim = c(0,1)
) +
geom_errorbar(
  data = data.frame(x = 0),
  aes(x = x, ymin = y0(x), ymax = y1(x)),
  color = orange, size = 1.5, width = 0.07
) +
scale_x_continuous(
  "SAT score",
  lim = c(-1,1),
  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = c(800, 1000, 1200, 1400, 1600)
) +
scale_y_continuous(
  "Probability of graduating",
  lim = c(0,1),
  labels = scales::percent_format(accuracy = 1)
) +
theme_simple
```
---

Using real data, researchers have to estimate $\color{#e64173}{\mathop{E}\left[ \text{Y}_{1,i} \mid \text{SAT}_{i} \right]}$ and $\color{#9370DB}{\mathop{E}\left[ \text{Y}_{0,i} \mid \text{SAT}_{i} \right]}$.

```{r, srd-gen-data, include = F, dev = "svg", fig.height = 5.5}
set.seed(12345)
srd_df <- tibble(
  x = seq(-1, 1, 0.02),
  y = (x < 0) * y0(x) + (x >= 0) * y1(x) + rnorm(length(x), sd = 0.07)
) %>% mutate(
  y = between(y, 0, 1) * y + (y < 0) * 0 + (y > 1) * 1
) %>% filter(
  x != 0
)
```

```{r, save-gg-srd, echo = F, dev = "svg", fig.height = 5.5}
gg_srd <- ggplot() +
geom_vline(xintercept = 0, color = slate, size = 1, alpha = 0.1) +
stat_function(
  data = data.frame(x = c(-1,0)), aes(x),
  fun = y0, color = purple, size = 1, alpha = 0.15,
  linetype = "solid",
  xlim = c(-1,0)
) +
stat_function(
  data = data.frame(x = c(0,1)), aes(x),
  fun = y0, color = purple, size = 1, alpha = 0.15,
  linetype = "dotted",
  xlim = c(0,1)
) +
stat_function(
  data = data.frame(x = c(-1,0)), aes(x),
  fun = y1, color = red_pink, size = 1, alpha = 0.15,
  linetype = "dotted",
  xlim = c(-1,0)
) +
stat_function(
  data = data.frame(x = c(0,1)), aes(x),
  fun = y1, color = red_pink, size = 1, alpha = 0.15,
  linetype = "solid",
  xlim = c(0,1)
) +
scale_x_continuous(
  "SAT score",
  # lim = c(-1,1),
  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = c(800, 1000, 1200, 1400, 1600)
) +
scale_y_continuous(
  "Probability of graduating",
  lim = c(0,1),
  labels = scales::percent_format(accuracy = 1)
) +
geom_point(
  data = srd_df, aes(x = x, y = y, color = x >= 0),
  size = 2, alpha = 0.8
) +
scale_color_manual(values = c(purple, red_pink)) +
theme_simple +
theme(legend.position = "none")
```

```{r, s5, echo = F, cache = T, dev = "svg", fig.height = 5.5}
gg_srd
```

---
One way to estimate the .hi-orange[jump] is to estimate a regression on each side of the cutoff.

```{r, s6, echo = F, cache = T, dev='svg', fig.height = 5.5}
lm_tmp <- lm(y ~ x + I(x>0), data = srd_df)
lm_fun <- function(x) predict(lm_tmp, data.frame(x = x))
s6 <- gg_srd +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(-1,-0.02),
  color = purple,
  size = 1.5
) +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(0.02,1),
  color = red_pink,
  size = 1.5
)
s6
```
---
count: false
One way to estimate the .hi-orange[jump] is to estimate a regression on each side of the cutoff.

```{r, s7, echo = F, cache = T, dev = "svg", fig.height = 5.5}
s6 +
geom_errorbar(
  data = tibble(x = 0, y0 = lm_fun(-1e-3), y1 = lm_fun(1e-3)),
  aes(x = x, ymin = y0, ymax = y1),
  color = orange, size = 1.5, width = 0.07
)
```
---
Another way is to estimate regressions using only data closer to the cutoff.

```{r, s8, echo = F, cache = T, dev = "svg", fig.height = 5.5}
lm_tmp <- lm(y ~ x + I(x>0), data = srd_df %>% filter(abs(x) < 0.5))
lm_fun <- function(x) predict(lm_tmp, data.frame(x = x))
gg_srd +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(-0.5,-0.02),
  color = purple,
  size = 1.5
) +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(0.02,0.5),
  color = red_pink,
  size = 1.5
) +
geom_errorbar(
  data = tibble(x = 0, y0 = lm_fun(-1e-3), y1 = lm_fun(1e-3)),
  aes(x = x, ymin = y0, ymax = y1),
  color = orange, size = 1.5, width = 0.07
)
```
---
Different choices of samples and models can lead to different estimates of the treatment effect!

```{r, s9, echo = F, cache = T, dev = "svg", fig.height = 5.5}
lm_tmp <- lm(y ~ x * I(x>0), data = srd_df)
lm_fun <- function(x) predict(lm_tmp, data.frame(x = x))
gg_srd +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(-1,-0.02),
  color = purple,
  size = 1.5
) +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(0.02,1),
  color = red_pink,
  size = 1.5
) +
geom_errorbar(
  data = tibble(x = 0, y0 = lm_fun(-1e-3), y1 = lm_fun(1e-3)),
  aes(x = x, ymin = y0, ymax = y1),
  color = orange, size = 1.5, width = 0.07
)
```
---
Different choices of samples and models can lead to different estimates of the treatment effect!

```{r, s10, echo = F, cache = T, dev = "svg", fig.height = 5.5}
lm_tmp <- lm(y ~ x * I(x>0), data = srd_df %>% filter(abs(x) < 0.5))
lm_fun <- function(x) predict(lm_tmp, data.frame(x = x))
gg_srd +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(-0.5,-0.02),
  color = purple,
  size = 1.5
) +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(0.02,0.5),
  color = red_pink,
  size = 1.5
) +
geom_errorbar(
  data = tibble(x = 0, y0 = lm_fun(-1e-3), y1 = lm_fun(1e-3)),
  aes(x = x, ymin = y0, ymax = y1),
  color = orange, size = 1.5, width = 0.07
)
```
---
Different choices of samples and models can lead to different estimates of the treatment effect!

```{r, s11, echo = F, cache = T, dev = "svg", fig.height = 5.5}
lm_tmp <- lm(y ~ x * I(x>0), data = srd_df %>% filter(abs(x) < 0.25))
lm_fun <- function(x) predict(lm_tmp, data.frame(x = x))
gg_srd +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(-0.25,-0.02),
  color = purple,
  size = 1.5
) +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(0.02,0.25),
  color = red_pink,
  size = 1.5
) +
geom_errorbar(
  data = tibble(x = 0, y0 = lm_fun(-1e-3), y1 = lm_fun(1e-3)),
  aes(x = x, ymin = y0, ymax = y1),
  color = orange, size = 1.5, width = 0.07
)
```
---
Different choices of samples and models can lead to different estimates of the treatment effect!

```{r, s12, echo = F, cache = T, dev = "svg", fig.height = 5.5}
lm_tmp <- lm(y ~ poly(x,2) * I(x>0), data = srd_df %>% filter(abs(x) < 0.5))
lm_fun <- function(x) predict(lm_tmp, data.frame(x = x))
gg_srd +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(-0.5,-0.02),
  color = purple,
  size = 1.5
) +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(0.02,0.5),
  color = red_pink,
  size = 1.5
) +
geom_errorbar(
  data = tibble(x = 0, y0 = lm_fun(-1e-3), y1 = lm_fun(1e-3)),
  aes(x = x, ymin = y0, ymax = y1),
  color = orange, size = 1.5, width = 0.07
)
```
---
Different choices of samples and models can lead to different estimates of the treatment effect!

```{r, s13, echo = F, cache = T, dev = "svg", fig.height = 5.5}
lm_tmp <- loess(y ~ x * I(x>0), data = srd_df %>% filter(abs(x) < 0.5))
lm_fun <- function(x) predict(lm_tmp, data.frame(x = x))
gg_srd +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(-0.5,-0.02),
  color = purple,
  size = 1.5
) +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(0.02,0.5),
  color = red_pink,
  size = 1.5
) +
geom_errorbar(
  data = tibble(x = 0, y0 = lm_fun(-1e-3), y1 = lm_fun(1e-3)),
  aes(x = x, ymin = y0, ymax = y1),
  color = orange, size = 1.5, width = 0.07
)
```
---
Different choices of samples and models can lead to different estimates of the treatment effect!

```{r, s14, echo = F, cache = T, dev = "svg", fig.height = 5.5}
lm_tmp <- loess(y ~ x * I(x>0), data = srd_df)
lm_fun <- function(x) predict(lm_tmp, data.frame(x = x))
gg_srd +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(-1,-0.02),
  color = purple,
  size = 1.5
) +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(0.02,1),
  color = red_pink,
  size = 1.5
) +
geom_errorbar(
  data = tibble(x = 0, y0 = lm_fun(-1e-3), y1 = lm_fun(1e-3)),
  aes(x = x, ymin = y0, ymax = y1),
  color = orange, size = 1.5, width = 0.07
)
```
---
Some modeling choices can find an effect even if none exists!

```{r, s15a, echo = F, cache = T, dev = "svg", fig.height = 5.5}
# No effect
set.seed(12345)
null <- function(x) 0.95 / (1 + exp(-5 * x))
null_srd <- tibble(
  x = seq(-1, 1, 0.02),
  y = null(x) + rnorm(length(x), sd = 0.07)
) %>% filter(x != 0)
# Estimate models
lm_tmp <- lm(y ~ x + I(x>0), data = null_srd)
lm_fun <- function(x) predict(lm_tmp, data.frame(x = x))
loess_tmp <- loess(y ~ x * I(x>0), data = null_srd)
loess_fun <- function(x) predict(loess_tmp, data.frame(x = x))
# Figure
ggplot() +
geom_vline(xintercept = 0, color = slate, size = 1, alpha = 0.1) +
stat_function(
  data = data.frame(x = c(-1,1)), aes(x),
  fun = null, color = orange, size = 1, alpha = 0.4,
  linetype = "solid",
  xlim = c(-1,1)
) +
scale_x_continuous(
  "SAT score",
  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = c(800, 1000, 1200, 1400, 1600)
) +
scale_y_continuous(
  "Probability of graduating",
  lim = c(0,1),
  labels = scales::percent_format(accuracy = 1)
) +
scale_color_manual(values = c(purple, red_pink)) +
theme_simple +
theme(legend.position = "none")
```
---
count: false
Some modeling choices can find an effect even if none exists!

```{r, s15b, echo = F, cache = T, dev = "svg", fig.height = 5.5}
ggplot() +
geom_vline(xintercept = 0, color = slate, size = 1, alpha = 0.1) +
stat_function(
  data = data.frame(x = c(-1,1)), aes(x),
  fun = null, color = orange, size = 1, alpha = 0.4,
  linetype = "solid",
  xlim = c(-1,1)
) +
scale_x_continuous(
  "SAT score",
  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = c(800, 1000, 1200, 1400, 1600)
) +
scale_y_continuous(
  "Probability of graduating",
  lim = c(0,1),
  labels = scales::percent_format(accuracy = 1)
) +
geom_point(
  data = null_srd, aes(x = x, y = y, color = x >= 0),
  size = 2, alpha = 0.8
) +
scale_color_manual(values = c(purple, red_pink)) +
theme_simple +
theme(legend.position = "none")
```
---
count: false
Some modeling choices can find an effect even if none exists!

```{r, s15c, echo = F, cache = T, dev = "svg", fig.height = 5.5}
ggplot() +
geom_vline(xintercept = 0, color = slate, size = 1, alpha = 0.1) +
stat_function(
  data = data.frame(x = c(-1,1)), aes(x),
  fun = null, color = orange, size = 1, alpha = 0.4,
  linetype = "solid",
  xlim = c(-1,1)
) +
scale_x_continuous(
  "SAT score",
  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = c(800, 1000, 1200, 1400, 1600)
) +
scale_y_continuous(
  "Probability of graduating",
  lim = c(0,1),
  labels = scales::percent_format(accuracy = 1)
) +
geom_point(
  data = null_srd, aes(x = x, y = y, color = x >= 0),
  size = 2, alpha = 0.8
) +
scale_color_manual(values = c(purple, red_pink)) +
theme_simple +
theme(legend.position = "none") +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(-1,-0.02),
  color = purple,
  size = 1.5
) +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(0.02,1),
  color = red_pink,
  size = 1.5
)
```
---
count: false
Some modeling choices can find an effect even if none exists!

```{r, s15d, echo = F, cache = T, dev = "svg", fig.height = 5.5}
ggplot() +
geom_vline(xintercept = 0, color = slate, size = 1, alpha = 0.1) +
stat_function(
  data = data.frame(x = c(-1,1)), aes(x),
  fun = null, color = orange, size = 1, alpha = 0.4,
  linetype = "solid",
  xlim = c(-1,1)
) +
scale_x_continuous(
  "SAT score",
  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = c(800, 1000, 1200, 1400, 1600)
) +
scale_y_continuous(
  "Probability of graduating",
  lim = c(0,1),
  labels = scales::percent_format(accuracy = 1)
) +
geom_point(
  data = null_srd, aes(x = x, y = y, color = x >= 0),
  size = 2, alpha = 0.8
) +
scale_color_manual(values = c(purple, red_pink)) +
theme_simple +
theme(legend.position = "none") +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(-1,-0.02),
  color = purple,
  size = 1.5
) +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = lm_fun,
  xlim = c(0.02,1),
  color = red_pink,
  size = 1.5
) +
geom_errorbar(
  data = tibble(x = 0, y0 = lm_fun(-1e-3), y1 = lm_fun(1e-3)),
  aes(x = x, ymin = y0, ymax = y1),
  color = orange, size = 1.5, width = 0.07
)
```
---
count: false
Some modeling choices can find an effect even if none exists!

```{r, s15e, echo = F, cache = T, dev = "svg", fig.height = 5.5}
ggplot() +
geom_vline(xintercept = 0, color = slate, size = 1, alpha = 0.1) +
stat_function(
  data = data.frame(x = c(-1,1)), aes(x),
  fun = null, color = orange, size = 1, alpha = 0.4,
  linetype = "solid",
  xlim = c(-1,1)
) +
scale_x_continuous(
  "SAT score",
  breaks = c(-1, -0.5, 0, 0.5, 1),
  labels = c(800, 1000, 1200, 1400, 1600)
) +
scale_y_continuous(
  "Probability of graduating",
  lim = c(0,1),
  labels = scales::percent_format(accuracy = 1)
) +
geom_point(
  data = null_srd, aes(x = x, y = y, color = x >= 0),
  size = 2, alpha = 0.8
) +
scale_color_manual(values = c(purple, red_pink)) +
theme_simple +
theme(legend.position = "none") +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = loess_fun,
  xlim = c(-1,-0.02),
  color = purple,
  size = 1.5
) +
stat_function(
  data = data.frame(x = c(-1, 1)),
  aes(x = x),
  fun = loess_fun,
  xlim = c(0.02,1),
  color = red_pink,
  size = 1.5
) +
geom_errorbar(
  data = tibble(x = 0, y0 = loess_fun(-1e-3), y1 = loess_fun(1e-3)),
  aes(x = x, ymin = y0, ymax = y1),
  color = orange, size = 1.5, width = 0.07
)
```

---
layout: false
# Regression discontinuity

**Q:** When should we trust a regression discontinuity comparison? 

- When is the comparison *internally valid*?

--

**A:** When we believe that **treatment is the only thing that changes** (other than observed outcomes) at the cutoff.

1. We don't want to see evidence of people **bunching** on one side of the threshold.
    - This could mean that people are **manipulating the assignment variable** near the cutoff so that they get the treatment.
    - Example: cheating among students who anticipate being close to the cutoff as a way to increase their score just enough to get the scholarship.
2. We don't want to see a **"jump" in other variables** at the cutoff.
    - This would mean that people on one side of the cutoff are **no longer comparable** to people on the other side!

---
# Regression discontinuity

**Q:** How can we tell if the treatment actually has a causal effect on the outcome? 

--

**A:** The treatment has an effect if **all three** of the statements below are true.

1. We believe that the regression discontinuity comparison is **internally valid.**
2. We can see that the **outcome variable "jumps"** at the cutoff ***when we look at the raw data.***
3. The estimate of the "jump" is **precise enough** to conclude that the effect is statistically significant.
