---
title: "Labor Markets"
subtitle: "EC 350: Labor Economics"
author: "[Kyle Raze](https://kyleraze.com)"
date: "Spring 2021"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    self_contained: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
      ratio: "16:9"
---

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(ggthemes, xaringanExtra, ggdag, viridis, knitr, dslabs, gapminder, extrafont, Ecdat, wooldridge, tidyverse, magrittr, janitor, kableExtra)
# Define colors
red_pink <- "#e64173"
met_slate <- "#272822" # metropolis font color 
purple <- "#9370DB"
green <- "#007935"
light_green <- "#7DBA97"
orange <- "#FD5F00"
turquoise <- "#44C1C4"
# knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  #dpi = 300,
  #cache = T,
  warning = F,
  message = F
)  
theme_simple <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
theme_market <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.title.x = element_text(hjust = 1, size = 17),
  axis.title.y = element_text(hjust = 1, angle = 0, size = 17),
  # axis.text.x = element_text(size = 12),
  # axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
theme_gif <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 17),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)
wrapper <- function(x, ...) paste(strwrap(x, ...), collapse = "\n")
shift_axis <- function(p, y=0){
  g <- ggplotGrob(p)
  dummy <- data.frame(y=y)
  ax <- g[["grobs"]][g$layout$name == "axis-b"][[1]]
  p + annotation_custom(grid::grobTree(ax, vp = grid::viewport(y=1, height=sum(ax$height))), 
                        ymax=y, ymin=y) +
    geom_hline(aes(yintercept=y), data = dummy, size = 0.5, color = met_slate) +
    theme(axis.text.x = element_blank(), 
          axis.ticks.x = element_blank())
}

xaringanExtra::use_scribble()
```

# Discussion

In the early 2000s, the French government shortened the work week from 39 to 35 hours, arguing that there would be work left over for the unemployed. 

- **Q:** Would you expect employment to increase, as the French government predicted? Why or why not?

---
# Lump of labor fallacy

--

**Q:** What is the fallacy?

**A:** The notion that **the amount of work** to be done in an economy **is fixed**, with the implication that increasing productivity or the pool of qualified workers will necessarily reduce job opportunities.


--

**Q:** Why this unnecessarily pessimistic view of the world incorrect?

--

**A:** The amount of work to to be done isn't fixed! Not all economic interactions are zero-sum!

- Job losses from automation can support the creation of new jobs in other industries. 
- As the population increases, demand for goods/services increases, which then increases demand for labor.


---
class: inverse, middle

# Supply and demand

---
# Supply and demand

## **Assumptions**

We will continue to make all of the same assumptions as we did when we modeled the "dilemmas" faced by workers and employers.

- Perfect competition in input and output markets
- Homogeneous workers within a market
- Various assumptions about preferences and production technology
- Perfect information and complete contracts


--

**Q:** Do these assumptions actually describe any real-world markets?

--

**A:** Strictly speaking, probably not. But they do allow us to make testable predictions!

> All models are wrong, but some are useful. &nbsp;&nbsp; &mdash; [George Box](https://en.wikipedia.org/wiki/George_E._P._Box)?

---
# Labor supply

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
supply <- function(x) 1 + (4/5)*x

ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = seq(0, 10, 1)) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = seq(0, 10, 1)) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = supply, color = red_pink, size = 1) +
  annotate("label", label = "S", x = 9, y = supply(9), color = red_pink, family = "Fira Sans", size = 5)
```
]
.pull-right[
## .hi-pink[Market labor supply curve]

Shows the number of .pink[workers] who are willing to work at specified wages, other things being equal.

- Or, alternatively, the amount of time .pink[workers] are willing to work at specified wages, other things being equal.

.pink[Upward sloping:] As w increases, L.sub[S] increases.
]

---
# Labor demand

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
demand <- function(x) 10 - x

ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = seq(0, 10, 1)) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = seq(0, 10, 1)) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1) +
  annotate("label", label = "D", x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5)
```
]
.pull-right[
## .hi-purple[Market labor demand curve]

Shows the number of workers .purple[employers] are willing to hire at specified wages, other things being equal.

- Or, alternatively, the amount of labor .purple[employers] are willing to employ at specified wages, other things being equal.

.purple[Downward sloping:] As w increases, L.sub[D] decreases.
]

---
# Equilibrium

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
market <- ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = seq(0, 10, 1)) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = seq(0, 10, 1)) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = supply, color = red_pink, size = 1) +
  annotate("label", label = "S", x = 9, y = supply(9), color = red_pink, family = "Fira Sans", size = 5) +
  stat_function(fun = demand, color = purple, size = 1) +
  annotate("label", label = "D", x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5)
market
```
]
.pull-right[
## **Market clearing**

The quantity of labor supplied equals the quantity of labor demanded: L.sub[S] .mono[=] L.sub[D].

No tendency for real wages to change.
]

---
count: false
# Equilibrium

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
market + 
  geom_point(aes(x = 5, y = 5), color = met_slate, size = 2) +
  geom_segment(aes(x = 5, y = 0, xend = 5, yend = 5), linetype  = "dashed", color = met_slate) + 
  geom_segment(aes(x = 0, y = 5, xend = 5, yend = 5), linetype  = "dashed", color = met_slate)
```
]
.pull-right[
## **Market clearing**

The quantity of labor supplied equals the quantity of labor demanded: L.sub[S] .mono[=] L.sub[D].

No tendency for real wages to change.
]

---
# Disequilibrium

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
market + geom_hline(yintercept = 3, linetype  = "dashed", color = met_slate) 
```
]
.pull-right[
## **Labor shortage**

The quantity of labor supplied is less than the quantity of labor demanded: L.sub[S] .mono[<] L.sub[D].

Real wages tend to rise.
]

---
count: false
# Disequilibrium

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
market + geom_hline(yintercept = 3, linetype  = "dashed", color = met_slate) +
  geom_segment(aes(x = 7, y = 0, xend = 7, yend = 3), linetype  = "dashed", color = met_slate) + 
  geom_point(aes(x = 7, y = 3), color = met_slate, size = 2) +
  geom_segment(aes(x = 2.5, y = 0, xend = 2.5, yend = 3), linetype  = "dashed", color = met_slate) + 
  geom_point(aes(x = 2.5, y = 3), color = met_slate, size = 2)
```
]
.pull-right[
## **Labor shortage**

The quantity of labor supplied is less than the quantity of labor demanded: L.sub[S] .mono[<] L.sub[D].

Real wages tend to rise.
]

---
# Disequilibrium

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
market + geom_hline(yintercept = 7, linetype  = "dashed", color = met_slate) 
```
]
.pull-right[
## **Unemployment**

The quantity of labor supplied is greater than the quantity of labor demanded: L.sub[S] .mono[>] L.sub[D].

Real wages tend to fall.
]

---
count: false
# Disequilibrium

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
market + geom_hline(yintercept = 7, linetype  = "dashed", color = met_slate) +
  geom_segment(aes(x = 3, y = 0, xend = 3, yend = 7), linetype  = "dashed", color = met_slate) + 
  geom_point(aes(x = 3, y = 7), color = met_slate, size = 2) +
  geom_segment(aes(x = 7.5, y = 0, xend = 7.5, yend = 7), linetype  = "dashed", color = met_slate) + 
  geom_point(aes(x = 7.5, y = 7), color = met_slate, size = 2) 
```
]
.pull-right[
## **Unemployment**

The quantity of labor supplied is greater than the quantity of labor demanded: L.sub[S] .mono[>] L.sub[D].

Real wages tend to fall.
]

---
# Unemployment

## **Definitions**

$\text{E}$ .mono[=] number of employed individuals

--

$\text{U}$ .mono[=] number of unemployed individuals

- Being "unemployed" means that you 1) want a job and 2) are actively looking for work.


--

$\text{P}$ .mono[=] population of interest (*e.g.,* US residents between the ages of 24 and 64)


--

Labor force $\text{LF}$ .mono[=] total number of individuals who want a job: $$\text{LF} = \text{E} + \text{U}$$

--

Labor force participation rate $\text{LFPR}$ .mono[=] percentage of the population in the labor force: $$\text{LFPR} = \dfrac{\text{LF}}{\text{P}} \times 100$$

---
# Unemployment rate

## **Definition**

> The unemployment rate $\text{UR}$ measures the percentage of the labor force that is unemployed: $$\text{UR} = \dfrac{\text{U}}{\text{LF}} \times 100$$

--

## **Caveats**

The unemployment rate you see in the news can give a misleading impression about the true extent of unemployment!

- Does not include **discouraged workers** who want a job, but stopped looking for work
- Does not include **underemployed workers** who are working part-time, but want to work full-time


--

Changes in economic conditions can also affect both the numerator and the denominator!

---
# Unemployment

Macroeconomists distinguish between three main types of unemployment:

1. **Frictional:** Unemployment caused by job search (*e.g.,* you after graduation).

2. **Structural:** Unemployment caused by mismatch between jobs and workers (*e.g.,* layoffs from technological change).

3. **Cyclical:** Unemployment caused by variations in the business cycle (*e.g.,* recessions).

---
# Determinants of labor supply

**Q:** What determines market labor supply?.super[.hi-pink[<span>&#8224;</span>]]

.footnote[.super[.hi-pink[<span>&#8224;</span>]] *Supply* .mono[=] Entire relationship between *wage* and the *quantity of labor supplied.*]

--

- Wages in other labor markets
- Non-labor income
- Preferences over work and leisure
- Working conditions
- "Fringe" benefits
- Number of qualified workers


--

**Changing the determinants** of supply **shifts** the entire supply curve.

**Changing the wage** results in **movement along** the supply curve. 

---
# Changes in labor supply

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:10
demand <- function(x) 10 - x
supply <- function(x) 1 + x
supply_2 <- function(x) -1 + x

q_1 <- uniroot(function(x) demand(x) - supply(x), range(x))$root
p_1 <- demand(q_1)

q_2 <- uniroot(function(x) demand(x) - supply_2(x), range(x))$root
p_2 <- demand(q_2)

base_2 <- ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(q_1), labels = c(expression(L^"*"))) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(p_1), labels = c(expression(w^"*"))) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1) +
  stat_function(fun = supply, color = red_pink, size = 1) +
  geom_point(x = q_1, y = p_1, color = met_slate, size = 2) +
  geom_segment(x = 0, xend = q_1, y = p_1, yend = p_1, color = met_slate, linetype = "dashed") +
  geom_segment(x = q_1, xend = q_1, y = 0, yend = p_1, color = met_slate, linetype = "dashed") +
  annotate("label", label = expression(S[1]), x = 8, y = supply(8), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D), x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5)

base_2
```
]
.pull-right[
## .pink[Supply increase]

At every wage, more people are willing to work than before.

Equilibrium quantity of labor increases.

Equilibrium wage decreases.
]

---
count: false
# Changes in labor supply

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:10
demand <- function(x) 10 - x
supply <- function(x) 1 + x
supply_2 <- function(x) -1 + x

q_1 <- uniroot(function(x) demand(x) - supply(x), range(x))$root
p_1 <- demand(q_1)

q_2 <- uniroot(function(x) demand(x) - supply_2(x), range(x))$root
p_2 <- demand(q_2)

ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(q_2), labels = c(expression(L^"*"))) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(p_2), labels = c(expression(w^"*"))) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1) +
  stat_function(fun = supply, color = red_pink, size = 1, alpha = 0.5) +
  stat_function(fun = supply_2, color = red_pink, size = 1) +
  geom_point(x = q_1, y = p_1, color = met_slate, size = 2, alpha = 0.5) +
  geom_point(x = q_2, y = p_2, color = met_slate, size = 2) +
  geom_segment(x = 0, xend = q_1, y = p_1, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = q_1, xend = q_1, y = 0, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = 0, xend = q_2, y = p_2, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(x = q_2, xend = q_2, y = 0, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(aes(x = 6 + 0.2, y = 7, xend = 8 - 0.2, yend = 7), arrow = arrow(angle = 20, length = unit(0.2, "cm"), type = "closed"), color = met_slate) + 
  annotate("label", label = expression(S[1]), x = 8, y = supply(8), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(S[2]), x = 9, y = supply_2(9), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D), x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5)
```
]
.pull-right[
## .pink[Supply increase]

At every wage, more people are willing to work than before.

Equilibrium quantity of labor increases.

Equilibrium wage decreases.
]

---
# Changes in labor supply

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
base_2
```
]
.pull-right[
## .pink[Supply decrease]

At every wage, fewer people are willing to work than before.

Equilibrium quantity of labor decreases.

Equilibrium wage increases.
]

---
count: false
# Changes in labor supply

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:10
demand <- function(x) 10 - x
supply <- function(x) 1 + x
supply_2 <- function(x) 3 + x

q_1 <- uniroot(function(x) demand(x) - supply(x), range(x))$root
p_1 <- demand(q_1)

q_2 <- uniroot(function(x) demand(x) - supply_2(x), range(x))$root
p_2 <- demand(q_2)

ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(q_2), labels = c(expression(L^"*"))) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(p_2), labels = c(expression(w^"*"))) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1) +
  stat_function(fun = supply, color = red_pink, size = 1, alpha = 0.5) +
  stat_function(fun = supply_2, color = red_pink, size = 1) +
  geom_point(x = q_1, y = p_1, color = met_slate, size = 2, alpha = 0.5) +
  geom_point(x = q_2, y = p_2, color = met_slate, size = 2) +
  geom_segment(x = 0, xend = q_1, y = p_1, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = q_1, xend = q_1, y = 0, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = 0, xend = q_2, y = p_2, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(x = q_2, xend = q_2, y = 0, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(aes(x = 6 - 0.2, y = 7, xend = 4 + 0.2, yend = 7), arrow = arrow(angle = 20, length = unit(0.2, "cm"), type = "closed"), color = met_slate) + 
  annotate("label", label = expression(S[1]), x = 8, y = supply(8), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(S[2]), x = 7, y = supply_2(7), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D), x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5)
```
]
.pull-right[
## .pink[Supply decrease]

At every wage, fewer people are willing to work than before.

Equilibrium quantity of labor decreases.

Equilibrium wage increases.
]



---
# Determinants of labor demand

**Q:** What determines market labor demand?.super[.hi-pink[<span>&#8224;</span>]]

.footnote[.super[.hi-pink[<span>&#8224;</span>]] *Demand* .mono[=] Entire relationship between *wage* and the *quantity of labor demanded.*]

--


- Demand for the output good
- Productivity of capital, labor, and other inputs
- Prices of other inputs
- Number of employers


--

**Changing the determinants** of demand **shifts** the entire demand curve.

**Changing the wage** results in **movement along** the demand curve. 

---
# Changes in labor demand

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:10
demand <- function(x) 10 - x
demand_2 <- function(x) 12 - x
supply <- function(x) 1 + x

q_1 <- uniroot(function(x) demand(x) - supply(x), range(x))$root
p_1 <- demand(q_1)

q_2 <- uniroot(function(x) demand_2(x) - supply(x), range(x))$root
p_2 <- demand_2(q_2)

base_1 <- ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(q_1), labels = c(expression(L^"*"))) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(p_1), labels = c(expression(w^"*"))) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1) +
  stat_function(fun = supply, color = red_pink, size = 1) +
  geom_point(x = q_1, y = p_1, color = met_slate, size = 2) +
  geom_segment(x = 0, xend = q_1, y = p_1, yend = p_1, color = met_slate, linetype = "dashed") +
  geom_segment(x = q_1, xend = q_1, y = 0, yend = p_1, color = met_slate, linetype = "dashed") +
  annotate("label", label = expression(S), x = 8, y = supply(8), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[1]), x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5)

base_1
```
]
.pull-right[
## .purple[Demand increase]

At every wage, employers are willing to hire more workers than before.

Equilibrium quantity of labor increases.

Equilibrium wage increases.
]

---
count: false
# Changes in labor demand

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:10
demand <- function(x) 10 - x
demand_2 <- function(x) 12 - x
supply <- function(x) 1 + x

q_1 <- uniroot(function(x) demand(x) - supply(x), range(x))$root
p_1 <- demand(q_1)

q_2 <- uniroot(function(x) demand_2(x) - supply(x), range(x))$root
p_2 <- demand_2(q_2)

ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(q_2), labels = c(expression(L^"*"))) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(p_2), labels = c(expression(w^"*"))) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1, alpha = 0.5) +
  stat_function(fun = demand_2, color = purple, size = 1) +
  stat_function(fun = supply, color = red_pink, size = 1) +
  geom_point(x = q_1, y = p_1, color = met_slate, size = 2, alpha = 0.5) +
  geom_point(x = q_2, y = p_2, color = met_slate, size = 2) +
  geom_segment(x = 0, xend = q_1, y = p_1, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = q_1, xend = q_1, y = 0, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = 0, xend = q_2, y = p_2, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(x = q_2, xend = q_2, y = 0, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(aes(x = 6 + 0.2, y = 4, xend = 8 - 0.2, yend = 4), arrow = arrow(angle = 20, length = unit(0.2,"cm"), type = "closed"), color = met_slate) + 
  annotate("label", label = expression(S), x = 8, y = supply(8), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[1]), x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[2]), x = 9, y = demand_2(9), color = purple, family = "Fira Sans", size = 5)
```
]
.pull-right[
## .purple[Demand increase]

At every wage, employers are willing to hire more workers than before.

Equilibrium quantity of labor increases.

Equilibrium wage increases.
]

---
# Changes in labor demand

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
base_1
```
]
.pull-right[
## .purple[Demand decrease]

At every wage, employers are willing to hire fewer workers than before.

Equilibrium quantity of labor decreases.

Equilibrium wage decreases.
]

---
count: false
# Changes in labor demand

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:10
demand <- function(x) 10 - x
demand_2 <- function(x) 8 - x
supply <- function(x) 1 + x

q_1 <- uniroot(function(x) demand(x) - supply(x), range(x))$root
p_1 <- demand(q_1)

q_2 <- uniroot(function(x) demand_2(x) - supply(x), range(x))$root
p_2 <- demand_2(q_2)

ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(q_2), labels = c(expression(L^"*"))) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(p_2), labels = c(expression(w^"*"))) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1, alpha = 0.5) +
  stat_function(fun = demand_2, color = purple, size = 1) +
  stat_function(fun = supply, color = red_pink, size = 1) +
  geom_point(x = q_1, y = p_1, color = met_slate, size = 2, alpha = 0.5) +
  geom_point(x = q_2, y = p_2, color = met_slate, size = 2) +
  geom_segment(x = 0, xend = q_1, y = p_1, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = q_1, xend = q_1, y = 0, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = 0, xend = q_2, y = p_2, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(x = q_2, xend = q_2, y = 0, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(aes(x = 6 - 0.2, y = 4, xend = 4 + 0.2, yend = 4), arrow = arrow(angle = 20, length = unit(0.2,"cm"), type = "closed"), color = met_slate) + 
  annotate("label", label = expression(S), x = 8, y = supply(8), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[1]), x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[2]), x = 7, y = demand_2(7), color = purple, family = "Fira Sans", size = 5)
```
]
.pull-right[
## .purple[Demand decrease]

At every wage, employers are willing to hire fewer workers than before.

Equilibrium quantity of labor decreases.

Equilibrium wage decreases.
]

---
class: clear-slide

**Q.sub[1]:** How would a construction boom in downtown Portland affect the market for welders in Portland?

**Q.sub[2]:** How would a construction boom in downtown Portland affect the market for welders *in Eugene*?

---
class: inverse, middle

# Immigration

---
class: clear-slide

**Q:** How does immigration affect equilibrium wages and employment?

- How might we model the effects of immigration? What determinants of labor supply or demand change?


--

**A.sub[1]:** If immigrant workers and native workers are **complements**, then they **do not compete in the same market.**

--

- An increase in immigrant labor would make native workers more productive .mono[-->] demand for native workers would increase!


--

**A.sub[2]:** If immigrant workers and native workers are **perfect substitutes**, then they **compete in the same market.**

- Immigration would increase labor supply.

--


- However, **immigrant workers are also consumers** .mono[-->] demand for goods and services would increase .mono[-->] labor demand would increase!

---
class: clear-slide

**Case 1:** Immigrant workers and native workers are complements.

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:10
demand <- function(x) 10 - x
demand_2 <- function(x) 12 - x
supply <- function(x) 3 + x

q_1 <- uniroot(function(x) demand(x) - supply(x), range(x))$root
p_1 <- demand(q_1)

q_2 <- uniroot(function(x) demand_2(x) - supply(x), range(x))$root
p_2 <- demand_2(q_2)

base_3 <- ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(q_1), labels = c(expression(L^"*"))) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(p_1), labels = c(expression(w^"*"))) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1) +
  stat_function(fun = supply, color = red_pink, size = 1) +
  geom_point(x = q_1, y = p_1, color = met_slate, size = 2) +
  geom_segment(x = 0, xend = q_1, y = p_1, yend = p_1, color = met_slate, linetype = "dashed") +
  geom_segment(x = q_1, xend = q_1, y = 0, yend = p_1, color = met_slate, linetype = "dashed") +
  annotate("label", label = expression(S[1]), x = 6, y = supply(6), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[1]), x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5)

base_3
```
]
.pull-right[

Complements .mono[-->] productivity of native workers increases .mono[-->] demand increases.


]

---
count: false
class: clear-slide

**Case 1:** Immigrant workers and native workers are complements.

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:10
demand <- function(x) 10 - x
demand_2 <- function(x) 12 - x
supply <- function(x) 3 + x

q_1 <- uniroot(function(x) demand(x) - supply(x), range(x))$root
p_1 <- demand(q_1)

q_2 <- uniroot(function(x) demand_2(x) - supply(x), range(x))$root
p_2 <- demand_2(q_2)

ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(q_2), labels = c(expression(L^"*"))) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(p_2), labels = c(expression(w^"*"))) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1, alpha = 0.5) +
  stat_function(fun = supply, color = red_pink, size = 1) +
  stat_function(fun = demand_2, color = purple, size = 1) +
  geom_point(x = q_1, y = p_1, color = met_slate, size = 2, alpha = 0.5) +
  geom_point(x = q_2, y = p_2, color = met_slate, size = 2) +
  geom_segment(x = 0, xend = q_1, y = p_1, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = q_1, xend = q_1, y = 0, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = 0, xend = q_2, y = p_2, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(x = q_2, xend = q_2, y = 0, yend = p_2, color = met_slate, linetype = "dashed") +
  
  geom_segment(aes(x = 6 + 0.2, y = 4, xend = 8 - 0.2, yend = 4), arrow = arrow(angle = 20, length = unit(0.2,"cm"), type = "closed"), color = met_slate) + 

  annotate("label", label = expression(S[1]), x = 6, y = supply(6), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[1]), x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[2]), x = 9, y = demand_2(9), color = purple, family = "Fira Sans", size = 5)
```
]
.pull-right[

Complements .mono[-->] productivity of native workers increases .mono[-->] demand increases.

- Equilibrium employment of native workers increases.
- Equilibrium wages increase.

]

---
class: clear-slide

**Case 2:** Immigrant workers and native workers are perfect substitutes.

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
base_3
```
]
.pull-right[

**Supply shift .mono[>] demand shift**

]

---
count: false
class: clear-slide

**Case 2:** Immigrant workers and native workers are perfect substitutes.

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:10
demand <- function(x) 10 - x
demand_2 <- function(x) 12 - x
supply <- function(x) 3 + x
supply_2 <- function(x) x

q_1 <- uniroot(function(x) demand(x) - supply(x), range(x))$root
p_1 <- demand(q_1)

q_2 <- uniroot(function(x) demand_2(x) - supply_2(x), range(x))$root
p_2 <- demand_2(q_2)

ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(q_2), labels = c(expression(L^"*"))) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(p_2), labels = c(expression(w^"*"))) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1, alpha = 0.5) +
  stat_function(fun = supply, color = red_pink, size = 1, alpha = 0.5) +
  stat_function(fun = demand_2, color = purple, size = 1) +
  stat_function(fun = supply_2, color = red_pink, size = 1) +
  geom_point(x = q_1, y = p_1, color = met_slate, size = 2, alpha = 0.5) +
  geom_point(x = q_2, y = p_2, color = met_slate, size = 2) +
  geom_segment(x = 0, xend = q_1, y = p_1, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = q_1, xend = q_1, y = 0, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = 0, xend = q_2, y = p_2, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(x = q_2, xend = q_2, y = 0, yend = p_2, color = met_slate, linetype = "dashed") +
  
  geom_segment(aes(x = 6 + 0.2, y = 4, xend = 8 - 0.2, yend = 4), arrow = arrow(angle = 20, length = unit(0.2,"cm"), type = "closed"), color = met_slate) + 
  geom_segment(aes(x = 5 + 0.2, y = 8, xend = 8 - 0.2, yend = 8), arrow = arrow(angle = 20, length = unit(0.2,"cm"), type = "closed"), color = met_slate) + 
  
  annotate("label", label = expression(S[1]), x = 6, y = supply(6), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[1]), x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(S[2]), x = 9, y = supply_2(9), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[2]), x = 9, y = demand_2(9), color = purple, family = "Fira Sans", size = 5)
```
]
.pull-right[

**Supply shift .mono[>] demand shift**

- Equilibrium employment increases.
- Equilibrium wages decrease.

]

---
class: clear-slide

**Case 2:** Immigrant workers and native workers are perfect substitutes.

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
base_3
```
]
.pull-right[

**Supply shift .mono[=] demand shift**
]

---
count: false
class: clear-slide

**Case 2:** Immigrant workers and native workers are perfect substitutes.

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:10
demand <- function(x) 10 - x
demand_2 <- function(x) 12 - x
supply <- function(x) 3 + x
supply_2 <- function(x) 1 + x

q_1 <- uniroot(function(x) demand(x) - supply(x), range(x))$root
p_1 <- demand(q_1)

q_2 <- uniroot(function(x) demand_2(x) - supply_2(x), range(x))$root
p_2 <- demand_2(q_2)

ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(q_2), labels = c(expression(L^"*"))) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(p_2), labels = c(expression(w^"*"))) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1, alpha = 0.5) +
  stat_function(fun = supply, color = red_pink, size = 1, alpha = 0.5) +
  stat_function(fun = demand_2, color = purple, size = 1) +
  stat_function(fun = supply_2, color = red_pink, size = 1) +
  geom_point(x = q_1, y = p_1, color = met_slate, size = 2, alpha = 0.5) +
  geom_point(x = q_2, y = p_2, color = met_slate, size = 2) +
  geom_segment(x = 0, xend = q_1, y = p_1, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = q_1, xend = q_1, y = 0, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = 0, xend = q_2, y = p_2, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(x = q_2, xend = q_2, y = 0, yend = p_2, color = met_slate, linetype = "dashed") +
  
  geom_segment(aes(x = 5.5 + 0.2, y = 4.5, xend = 7.5 - 0.2, yend = 4.5), arrow = arrow(angle = 20, length = unit(0.2,"cm"), type = "closed"), color = met_slate) + 
  geom_segment(aes(x = 5 + 0.2, y = 8, xend = 7 - 0.2, yend = 8), arrow = arrow(angle = 20, length = unit(0.2,"cm"), type = "closed"), color = met_slate) + 
  
  annotate("label", label = expression(S[1]), x = 6, y = supply(6), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[1]), x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(S[2]), x = 8, y = supply_2(8), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[2]), x = 9, y = demand_2(9), color = purple, family = "Fira Sans", size = 5)
```
]
.pull-right[

**Supply shift .mono[=] demand shift**

- Equilibrium employment increases.
- Equilibrium wages remain constant.

]

---
class: clear-slide

**Case 2:** Immigrant workers and native workers are perfect substitutes.

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
base_3
```
]
.pull-right[

**Supply shift .mono[<] demand shift**

]


---
count: false
class: clear-slide

**Case 2:** Immigrant workers and native workers are perfect substitutes.

.pull-left[
```{r, echo = FALSE, fig.height = 5.5, fig.width = 5.5, dev = "svg"}
x <- 0:10
demand <- function(x) 10 - x
demand_2 <- function(x) 13 - x
supply <- function(x) 3 + x
supply_2 <- function(x) 2 + x

q_1 <- uniroot(function(x) demand(x) - supply(x), range(x))$root
p_1 <- demand(q_1)

q_2 <- uniroot(function(x) demand_2(x) - supply_2(x), range(x))$root
p_2 <- demand_2(q_2)

ggplot(data = data.frame(x = 0), mapping = aes(x = x)) +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(q_2), labels = c(expression(L^"*"))) +
  scale_y_continuous(limits = c(0, 10.5), expand = c(0, 0), breaks = c(p_2), labels = c(expression(w^"*"))) +
  theme_market +
  labs(x = "L", y = "w") +
  stat_function(fun = demand, color = purple, size = 1, alpha = 0.5) +
  stat_function(fun = supply, color = red_pink, size = 1, alpha = 0.5) +
  stat_function(fun = demand_2, color = purple, size = 1) +
  stat_function(fun = supply_2, color = red_pink, size = 1) +
  geom_point(x = q_1, y = p_1, color = met_slate, size = 2, alpha = 0.5) +
  geom_point(x = q_2, y = p_2, color = met_slate, size = 2) +
  geom_segment(x = 0, xend = q_1, y = p_1, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = q_1, xend = q_1, y = 0, yend = p_1, color = met_slate, linetype = "dashed", alpha = 0.5) +
  geom_segment(x = 0, xend = q_2, y = p_2, yend = p_2, color = met_slate, linetype = "dashed") +
  geom_segment(x = q_2, xend = q_2, y = 0, yend = p_2, color = met_slate, linetype = "dashed") +
  
  geom_segment(aes(x = 5.5 + 0.2, y = 4.5, xend = 8.5 - 0.2, yend = 4.5), arrow = arrow(angle = 20, length = unit(0.2,"cm"), type = "closed"), color = met_slate) + 
  geom_segment(aes(x = 5 + 0.2, y = 8, xend = 6 - 0.2, yend = 8), arrow = arrow(angle = 20, length = unit(0.2,"cm"), type = "closed"), color = met_slate) + 

  annotate("label", label = expression(S[1]), x = 6, y = supply(6), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[1]), x = 9, y = demand(9), color = purple, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(S[2]), x = 7, y = supply_2(7), color = red_pink, family = "Fira Sans", size = 5) +
  annotate("label", label = expression(D[2]), x = 9, y = demand_2(9), color = purple, family = "Fira Sans", size = 5)
```
]
.pull-right[

**Supply shift .mono[<] demand shift**

- Equilibrium employment increases.
- Equilibrium wages increase.

]

---
# Immigration

**Q:** How does immigration affect equilibrium wages and employment?

--

**A:** Employment will increase, but the effect of immigration on wages is theoretically ambiguous.

--

- Whether immigration increases or decreases wages is an **empirical question!**


---
# Housekeeping

**Assigned reading for Wednesday:** [The Economic Impact of Migrants from Hurricane Maria](https://www.nber.org/papers/w27718) by Giovanni Peri, Derek Rury, and Justin C. Wiltshire (2020).

- I will post the paper and quiz tonight!
- Read the non-technical sections and the first two figures (details in quiz instructions).
- Reading Quiz 5 is due by **Wednesday, May 5th at 16:00**. 





